---
title: "Crude Oil Production and Evacuation Analytics: Drivers of Zone-Level Throughput Performance"
author: "Abiba Badmus — 2026/31/E/1647"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    self-contained: true
    fig-cap-location: bottom
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
# ── Python configuration ─────────────────────────────────────────────────────
# Uses an environment variable (QUARTO_PYTHON) if set; otherwise falls back to
# the default system Python so the document is reproducible on any machine.
if (requireNamespace("reticulate", quietly = TRUE)) {
  env_python <- Sys.getenv("QUARTO_PYTHON", unset = "")
  if (nzchar(env_python) && file.exists(env_python)) {
    reticulate::use_python(env_python, required = FALSE)
  }
  # If env var not set, reticulate picks up the system/conda Python automatically
}

# ── R libraries ───────────────────────────────────────────────────────────────
pkgs_core <- c("readr", "dplyr", "tidyr", "tibble", "purrr", "forcats",
               "ggplot2", "lubridate", "scales", "patchwork", "broom", "knitr",
               "sandwich", "lmtest")
invisible(lapply(pkgs_core, library, character.only = TRUE))

# kableExtra: use shim if system dependency unavailable
if (!requireNamespace("kableExtra", quietly = TRUE)) {
  kable_styling <- function(x, ...) x
} else {
  library(kableExtra)
}

knitr::opts_chunk$set(fig.width = 9, fig.height = 5, dpi = 150)

# ── Minimal ggcorrplot replacement using ggplot2 ──────────────────────────────
# Avoids the stringi/harfbuzz system-dependency issue while producing a
# publication-quality correlation heatmap with annotated r-values.
gg_corr_tile <- function(cor_mat, title = "Pearson Correlation Matrix") {
  df_long <- as.data.frame(as.table(cor_mat))
  names(df_long) <- c("Var1", "Var2", "r")
  # Mask upper triangle (keep lower + diagonal)
  mat_order <- rownames(cor_mat)
  df_long$row_idx <- match(df_long$Var1, mat_order)
  df_long$col_idx <- match(df_long$Var2, mat_order)
  df_long <- df_long[df_long$row_idx >= df_long$col_idx, ]
  ggplot(df_long, aes(x = Var2, y = Var1, fill = r)) +
    geom_tile(colour = "white", linewidth = 0.5) +
    geom_text(aes(label = round(r, 2)), size = 3.2, colour = "black") +
    scale_fill_gradient2(low = "#d73027", mid = "white", high = "#1a9850",
                         midpoint = 0, limits = c(-1, 1), name = "r") +
    scale_x_discrete(limits = mat_order) +
    scale_y_discrete(limits = rev(mat_order)) +
    labs(title = title, x = NULL, y = NULL) +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 40, hjust = 1),
          panel.grid = element_blank())
}
```

```{python setup-py, include=FALSE}
# ── Python libraries ───────────────────────────────────────────────────────────
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
import matplotlib.ticker as mticker
import seaborn as sns
from scipy import stats
import statsmodels.formula.api as smf
import statsmodels.api as sm
from IPython.display import display

sns.set_theme(style="whitegrid", palette="muted")
plt.rcParams["figure.dpi"] = 150
```

```{r load-r, include=FALSE}
# ── R: Load and clean data ─────────────────────────────────────────────────────
parse_num <- function(x) as.numeric(gsub("[^0-9.]", "", trimws(x)))

pick_data_path <- function(file_name) {
  candidates <- c(
    file.path("data", file_name),
    file.path("Abiba Badmus", "data", file_name),
    file.path("..", "Abiba Badmus", "data", file_name)
  )
  hit <- candidates[file.exists(candidates)]
  if (length(hit) > 0) hit[1] else candidates[1]
}

prod_r <- read_csv(pick_data_path("daily production report.csv"),
                   col_types = cols(.default = col_character())) |>
  mutate(
    DATE          = dmy(DATE),
    BUDGET        = parse_num(BUDGET),
    ACTUAL        = ifelse(trimws(ACTUAL) %in% c("-", "--", "  -   ", ""), "0", ACTUAL),
    ACTUAL        = parse_num(ACTUAL),
    ZONE          = trimws(ZONE),
    FLOW_STATIONS = trimws(FLOW_STATIONS),
    shortfall     = ACTUAL - BUDGET
  )

evac_r <- read_csv(pick_data_path("evacuation_report.csv"),
                   col_types = cols(.default = col_character())) |>
  mutate(
    DATE                   = dmy(DATE),
    evacuation_capacity    = parse_num(evacuation_capacity),
    evacuation_utilization = parse_num(evacuation_utilization),
    ZONE                   = trimws(ZONE),
    utilization_rate       = ifelse(evacuation_capacity > 0,
                                    evacuation_utilization / evacuation_capacity,
                                    NA_real_)
  )

# Zone-level aggregated production
prod_zone_r <- prod_r |>
  group_by(DATE, ZONE) |>
  summarise(BUDGET_zone    = sum(BUDGET,    na.rm = TRUE),
            ACTUAL_zone    = sum(ACTUAL,    na.rm = TRUE),
            shortfall_zone = sum(shortfall, na.rm = TRUE),
            .groups = "drop")

# Master joined dataset
df_r <- prod_zone_r |>
  left_join(evac_r, by = c("DATE", "ZONE")) |>
  mutate(period = case_when(
    month(DATE) <= 2 ~ "Jan-Feb",
    month(DATE) <= 4 ~ "Mar-Apr",
    TRUE             ~ "May+"
  ))

cat("R merge check -> rows:", nrow(df_r),
    "| missing evacuation capacity:", sum(is.na(df_r$evacuation_capacity)), "\n")
```

```{python load-py, include=FALSE}
# ── Python: Load and clean data ────────────────────────────────────────────────
from pathlib import Path

def parse_num(s):
    return pd.to_numeric(
        s.astype(str).str.strip().str.replace(r"[^\d.]", "", regex=True),
        errors="coerce"
    )

def pick_data_path(file_name: str) -> str:
  candidates = [
    Path("data") / file_name,
    Path("Abiba Badmus") / "data" / file_name,
    Path("..") / "Abiba Badmus" / "data" / file_name,
  ]
  for p in candidates:
    if p.exists():
      return str(p)
  return str(candidates[0])

prod_py = pd.read_csv(pick_data_path("daily production report.csv"), dtype=str)
prod_py["DATE"]          = pd.to_datetime(prod_py["DATE"].str.strip(), dayfirst=True)
prod_py["BUDGET"]        = parse_num(prod_py["BUDGET"])
prod_py["ACTUAL"]        = prod_py["ACTUAL"].str.strip().replace({"-": "0", "  -   ": "0", "": "0"})
prod_py["ACTUAL"]        = parse_num(prod_py["ACTUAL"]).fillna(0)
prod_py["ZONE"]          = prod_py["ZONE"].str.strip()
prod_py["FLOW_STATIONS"] = prod_py["FLOW_STATIONS"].str.strip()
prod_py["shortfall"]     = prod_py["ACTUAL"] - prod_py["BUDGET"]

evac_py = pd.read_csv(pick_data_path("evacuation_report.csv"), dtype=str)
evac_py["DATE"]                   = pd.to_datetime(evac_py["DATE"].str.strip(), dayfirst=True)
evac_py["evacuation_capacity"]    = parse_num(evac_py["evacuation_capacity"])
evac_py["evacuation_utilization"] = parse_num(evac_py["evacuation_utilization"])
evac_py["ZONE"]                   = evac_py["ZONE"].str.strip()
evac_py["utilization_rate"]       = evac_py["evacuation_utilization"] / evac_py["evacuation_capacity"].replace(0, np.nan)

prod_zone_py = prod_py.groupby(["DATE", "ZONE"]).agg(
    BUDGET_zone=("BUDGET",    "sum"),
    ACTUAL_zone=("ACTUAL",    "sum"),
    shortfall_zone=("shortfall", "sum")
).reset_index()

df_py = prod_zone_py.merge(evac_py, on=["DATE", "ZONE"], how="left")
print(f"Python merge check -> rows: {len(df_py)} | missing evacuation capacity: {df_py['evacuation_capacity'].isna().sum()}")
```

---

## Executive Summary

This analysis identifies where daily crude oil production losses are occurring across OML 18 and quantifies which operational levers can most effectively close the gap against plan. The analysis covers two internal datasets extracted from OML 18 systems (January 1 – May 5, 2026): 625 station-level production records across four active flow stations in two evacuation zones, and 250 zone-level evacuation capacity records. Five techniques — EDA, visualization, hypothesis testing, correlation, and regression — are applied in sequence to build from description to quantified recommendation.

Findings show persistent, system-wide underperformance: 86.4% of station-days fall below budget. Zone differences are directionally present but not statistically robust (Kruskal-Wallis p = 0.195); station-level differences, however, are highly significant (ANOVA F = 32.6, p < 0.001), with CC2 the deepest underperformer. Correlation and regression both identify evacuation capacity as the primary modifiable lever (r = −0.36 with shortfall; regression coefficient = −0.56 bbl/day shortfall per 1 bbl/day capacity, p < 0.001). The model explains 29% of shortfall variance, so recommendations are expressed as conditional operational projections, not causal guarantees.

**Recommendation:** A targeted 10,000 bbl/day fleet expansion on Zone 1 alternative evacuation routes is associated with an expected recovery of approximately 5,602 bbl/day under current operating conditions. The uncertainty range and scenario bounds are reported in Technique 5 to support risk-aware implementation.

*(Word count: ~190)*

---

## Professional Disclosure

**Job title:** Head, Alternative Crude Oil Evacuation
**Organisation sector:** Upstream oil and gas operations — national crude production, zone-level throughput management, and evacuation logistics

### Technique 1 — Exploratory Data Analysis

As Head of Alternative Crude Oil Evacuation, every working day begins with a review of the previous day's production figures: did each flow station meet its `BUDGET` target, and what is the gap in `ACTUAL`? Before any statistical test can be run, I need to understand the distributional shape of that gap across 625 daily records spanning four flow stations (ALK, CC1, CC2, CC3) and two zones. Are shortfalls concentrated in certain stations or spread evenly? Are there extreme outlier days — caused by pipeline maintenance or force majeure — that would distort any zone-level comparison? EDA answers these questions first and ensures that every downstream test and model rests on verified, understood data rather than unexamined assumptions. This mirrors the operational health-check I conduct before presenting production performance to senior management.

**Alternative considered:** Jumping directly to regression on the combined dataset. Rejected because undetected outliers in `ACTUAL` (including blank/zero values in early January) would silently bias regression coefficients and produce misleading conclusions about which zones are genuinely underperforming.

**Limitation:** EDA is descriptive only. A zone that visually appears to underperform may still fail to do so significantly once sample variance is accounted for. Formal tests are required before drawing operational conclusions.

### Technique 2 — Data Visualization

Upstream operations leadership does not make decisions from tables of numbers. The briefings I prepare for management are built around time-series charts of daily `ACTUAL` vs `BUDGET` production, zone-level comparison charts, and capacity utilization trend lines drawn from `evacuation_capacity` and `evacuation_utilization`. The `DATE` field enables a full January–May 2026 timeline view, while `ZONE` and `FLOW_STATIONS` provide the categorical groupings needed for comparison. Visualization allows me to spot bottleneck periods — weeks where Zone 1 (CC2/CC3) and Zone 2 (ALK/CC1) diverge — in a format that a non-technical executive can act on immediately. These charts are directly reusable in production performance briefings.

**Alternative considered:** Summary statistics tables only. Rejected because tables do not communicate temporal patterns or cross-zone divergence, precisely the patterns that matter most for evacuation scheduling decisions.

**Limitation:** Visual patterns can appear more decisive than they are. A chart showing Zone 2 consistently below budget could be driven by one outlier period. All visual findings are followed by formal hypothesis tests before policy action is recommended.

### Technique 3 — Hypothesis Testing

My organisation operates four flow stations across two zones, and a persistent management question is whether Zone 1 and Zone 2 perform statistically differently against their respective `BUDGET` targets, or whether the apparent difference is within normal operational variance. If the difference is statistically significant, that is a resource allocation issue: it justifies redirecting evacuation capacity or maintenance attention to the underperforming zone. If it is not significant, the variability is systemic and calls for a different intervention. A p-value and effect size turn what is currently an informal management opinion into a defensible, data-backed operational directive.

**Alternative considered:** Kruskal-Wallis non-parametric test as the primary method, since daily production data can exhibit skew. This is run as a robustness check alongside parametric ANOVA — as it turns out, the non-parametric result overrides the parametric one.

**Limitation:** After splitting by both `ZONE` and `FLOW_STATIONS`, some station-level subgroups may have fewer observations than ideal for high-power comparisons. Zone-level comparisons are adequately powered; station-level results are interpreted with explicit power caveats.

### Technique 4 — Correlation Analysis

Before including `evacuation_capacity` and `evacuation_utilization` as predictors in regression, I need to establish whether they are actually related to production performance. My operational hypothesis is that low utilization relative to capacity should correlate with better production attainment. Correlation analysis on the merged daily dataset confirms or refutes this and reveals which evacuation variables are worth including as predictors. It also guards against multicollinearity. Given the confirmed non-normality of shortfall distributions (Shapiro-Wilk p < 0.05), Spearman rank correlation is the primary non-parametric measure; Pearson is reported alongside for direct comparison.

**Alternative considered:** Entering all evacuation and production variables directly into regression without prior screening. Rejected because this produces inflated standard errors and misleading coefficient signs when predictors are correlated.

**Limitation:** Correlation measures linear (Pearson) or monotonic (Spearman) association only. The relationship between evacuation utilization and production attainment may be non-linear — utilization below 40% may have limited impact while utilization above 80% may sharply constrain output. This is explored visually before regression specification.

### Technique 5 — Multiple Regression

The commercial and strategic purpose of this analysis is a daily production attainment model: given a station's `BUDGET` target, its `ZONE`, its `evacuation_capacity`, and its `utilization_rate`, what `ACTUAL` production shortfall should be expected? A regression equation with `shortfall` as the outcome variable provides coefficient-level answers to the question every evacuation planning meeting asks: which factor, if improved by one unit, recovers the most daily barrels? Every coefficient in this model maps to a recoverable share of annual throughput targets.

**Alternative considered:** Random forest regression to capture non-linear interactions between zone, station, and capacity variables. Rejected because the current dataset is relatively small for stable ensemble generalisation and the output would be less interpretable for management briefing.

**Limitation:** The regression assumes approximately linear relationships between predictors and shortfall. If the true relationship is threshold-based — for example, output only suffers once capacity falls below a critical level — diagnostic plots will reveal this and a transformed specification can be applied.

---

## Data Collection & Sampling

| Attribute | Detail |
|-----------|--------|
| **Source system** | Internal daily production and evacuation reporting — OML 18 (Oil Mining Lease 18) Upstream Operations |
| **Zone structure** | Zone 1: CC2, CC3 · Zone 2: ALK, CC1. Three additional planned stations are non-operational and excluded |
| **Dataset 1** | `daily production report.csv` — 625 station-day records, flow-station budget vs actual production |
| **Dataset 2** | `evacuation_report.csv` — 250 zone-day records, evacuation capacity and utilization |
| **Time period** | 1 January – 5 May 2026 (125 calendar days; May 2026 partial — 5 days only) |
| **Collection method** | Direct extraction from internal operational reporting system; author holds system access as accountability holder |
| **Sample frame** | Full operational population — no sampling; all active flow stations and zones included |
| **Selection bias** | None deliberate. Zero/blank `ACTUAL` values in early January treated as non-operational periods, not collection errors |
| **Anonymisation** | Flow-station codes and zone designators are internal operational names; no employee or customer PII present |
| **Ethical clearance** | Internal line management approval and data-governance clearance obtained for academic use of operational aggregates |
| **Restrictions** | Company-confidential absolute volume figures anonymised; relative ratios and shortfall statistics are publishable under internal approval |

### Data Governance, Extraction Workflow, and Sampling Justification

- **System of record and extraction pipeline:** Daily production and evacuation records were exported from internal OML 18 operational reporting modules and saved as controlled CSV snapshots (`daily production report.csv`, `evacuation_report.csv`) used in this analysis.
- **Extraction governance:** The extraction event, approver identity, and approval reference are logged in the internal operations audit trail maintained by the line function; only aggregate operational fields were exported for academic use.
- **Privacy and confidentiality:** No direct personal identifiers or employee-level fields are present. Flow-station and zone labels are operational asset codes; absolute sensitive commercial values are not disclosed outside approved internal channels.
- **Sampling frame rationale:** This is a full operational census of active assets over the covered period (Jan 1-May 5, 2026), not a sampled subset. For zone-level comparisons, the effective group sizes are sufficient for stable inference on medium effects, while the observed zone effect (eta-squared = 0.007) is substantively negligible regardless of statistical significance under alternative assumptions.
- **Defence-ready caveat:** Because May covers only 5 days, formal trend inference is reported both with and without partial-May observations to avoid overstating deterioration.

---

## Data Description

### Variable Dictionary

| Variable | Source | Type | Unit | Description |
|----------|--------|------|------|-------------|
| `DATE` | Both | Date | YYYY-MM-DD | Calendar date of observation |
| `ZONE` | Both | Categorical | — | Evacuation zone (Zone 1 / Zone 2) |
| `FLOW_STATIONS` | Production | Categorical | — | Flow station code: ALK, CC1, CC2, CC3 |
| `BUDGET` | Production | Numeric | bbl/day | Planned production target for the station-day |
| `ACTUAL` | Production | Numeric | bbl/day | Realised production for the station-day |
| `shortfall` | Derived | Numeric | bbl/day | `ACTUAL − BUDGET`; negative = underperformance |
| `evacuation_capacity` | Evacuation | Numeric | bbl/day | Maximum daily volume the alternative route can carry |
| `evacuation_utilization` | Evacuation | Numeric | bbl/day | Actual daily volume moved via alternative route |
| `utilization_rate` | Derived | Numeric | ratio | `evacuation_utilization / evacuation_capacity` |
| `BUDGET_zone` | Derived | Numeric | bbl/day | Sum of station budgets within zone-day |
| `ACTUAL_zone` | Derived | Numeric | bbl/day | Sum of station actuals within zone-day |
| `shortfall_zone` | Derived | Numeric | bbl/day | Zone-day shortfall = `ACTUAL_zone − BUDGET_zone` |

::: {.panel-tabset}

## R

```{r eda-describe-r}
cat("Production records:", nrow(prod_r), "rows |", ncol(prod_r), "columns\n")
cat("Evacuation records:", nrow(evac_r), "rows |", ncol(evac_r), "columns\n")
cat("Merged zone dataset:", nrow(df_r), "rows\n\n")

prod_r |>
  select(BUDGET, ACTUAL, shortfall) |>
  summary() |>
  kable(caption = "Production dataset — key numeric summaries") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r missing-r}
prod_r |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "Column", values_to = "NA_count") |>
  filter(NA_count > 0) |>
  kable(caption = "Missing values in production dataset") |>
  kable_styling(full_width = FALSE)

cat("\nNote: ACTUAL = 0 records (non-operational days):",
    sum(prod_r$ACTUAL == 0), "\n")

prod_r |>
  count(ZONE, FLOW_STATIONS) |>
  kable(caption = "Record counts by Zone and Flow Station") |>
  kable_styling(full_width = FALSE)
```

## Python

```{python eda-describe-py}
print(f"Production records: {prod_py.shape[0]} rows | {prod_py.shape[1]} columns")
print(f"Evacuation records: {evac_py.shape[0]} rows | {evac_py.shape[1]} columns")
print(f"Merged zone dataset: {df_py.shape[0]} rows\n")

display(prod_py[["BUDGET", "ACTUAL", "shortfall"]].describe().round(2))
```

```{python missing-py}
na_counts = prod_py.isnull().sum()
na_counts = na_counts[na_counts > 0]
if len(na_counts):
    print("Missing values:\n", na_counts)
else:
    print("No missing values after cleaning.\n")

print(f"ACTUAL = 0 records (non-operational days): {(prod_py['ACTUAL'] == 0).sum()}")
print("\nRecord counts by Zone and Flow Station:")
display(prod_py.groupby(["ZONE", "FLOW_STATIONS"]).size().reset_index(name="count"))
```

:::

---

## Technique 1 — Exploratory Data Analysis

**Theory recap:** EDA (Tukey, 1977) is a philosophy of letting data guide analysis before imposing assumptions. Core tools include distributional summaries, outlier detection via the IQR method, and skewness/kurtosis statistics. Anscombe's Quartet (1973) demonstrates why summary statistics alone are insufficient: four datasets with identical means and standard deviations can have radically different visual structures. EDA mitigates this risk.

**Business justification:** Before any inferential test, this section establishes the shape of the production shortfall distribution, identifies extreme outlier days (maintenance events, force majeure), and reveals how performance varies across zones and flow stations. These findings directly inform which statistical tests are appropriate in later sections.

::: {.panel-tabset}

## R

```{r eda-anscombe-r}
# Anscombe's Quartet: identical summaries can hide different structures
ans_long <- anscombe |>
  tibble::as_tibble() |>
  mutate(obs = row_number()) |>
  pivot_longer(-obs, names_to = "var", values_to = "value") |>
  mutate(set = substr(var, 2, 2), axis = substr(var, 1, 1)) |>
  select(-var) |>
  pivot_wider(names_from = axis, values_from = value)

ans_stats <- ans_long |>
  group_by(set) |>
  summarise(
    mean_x = mean(x),
    mean_y = mean(y),
    cor_xy = cor(x, y),
    .groups = "drop"
  )

kable(ans_stats |> mutate(across(where(is.numeric), ~ round(., 3))),
      caption = "Anscombe's Quartet: nearly identical summary statistics") |>
  kable_styling(full_width = FALSE)

ggplot(ans_long, aes(x = x, y = y)) +
  geom_point(colour = "#1f77b4", alpha = 0.85) +
  geom_smooth(method = "lm", se = FALSE, colour = "#d62728", linewidth = 0.8) +
  facet_wrap(~set, ncol = 2) +
  labs(title = "Anscombe's Quartet: same summaries, different data geometry",
       x = "x", y = "y") +
  theme_minimal(base_size = 11)
```

```{r eda-r}
p1 <- ggplot(prod_r, aes(x = shortfall, fill = ZONE)) +
  geom_histogram(bins = 40, alpha = 0.7, position = "identity") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "red") +
  facet_wrap(~ZONE, ncol = 1) +
  scale_x_continuous(labels = comma) +
  labs(title = "Distribution of Daily Production Shortfall by Zone",
       subtitle = "Red dashed line = on-target (shortfall = 0). Left of line = underperformance.",
       x = "Shortfall (bbl/day)", y = "Count") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")

p2 <- ggplot(prod_r, aes(x = FLOW_STATIONS, y = shortfall, fill = ZONE)) +
  geom_boxplot(alpha = 0.7, outlier.colour = "red", outlier.size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_y_continuous(labels = comma) +
  labs(title = "Shortfall Distribution by Flow Station",
       x = "Flow Station", y = "Shortfall (bbl/day)") +
  theme_minimal(base_size = 11)

p1 / p2
```

```{r eda-outliers-r}
# ── Outlier detection via IQR fences ──────────────────────────────────────────
# Any shortfall beyond Q1 - 1.5*IQR or Q3 + 1.5*IQR is flagged as an outlier.
# Operationally these likely correspond to maintenance shutdowns or force-majeure.
outlier_summary <- prod_r |>
  group_by(FLOW_STATIONS) |>
  summarise(
    Q1       = quantile(shortfall, 0.25, na.rm = TRUE),
    Q3       = quantile(shortfall, 0.75, na.rm = TRUE),
    IQR      = Q3 - Q1,
    lower_fence = Q1 - 1.5 * IQR,
    upper_fence = Q3 + 1.5 * IQR,
    n_outliers  = sum(shortfall < Q1 - 1.5 * IQR | shortfall > Q3 + 1.5 * IQR,
                      na.rm = TRUE),
    pct_outliers = round(n_outliers / n() * 100, 1),
    .groups = "drop"
  )

kable(outlier_summary |> mutate(across(where(is.numeric), ~ round(., 0))),
      caption = "IQR-based outlier detection by flow station") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

cat("\nTotal IQR-flagged extreme observations:",
    sum(prod_r$shortfall < quantile(prod_r$shortfall, 0.25, na.rm=TRUE) - 
                            1.5 * IQR(prod_r$shortfall, na.rm=TRUE) |
        prod_r$shortfall > quantile(prod_r$shortfall, 0.75, na.rm=TRUE) + 
                            1.5 * IQR(prod_r$shortfall, na.rm=TRUE), na.rm=TRUE),
    "of", nrow(prod_r), "records\n")
cat("Decision: retain extreme low values as genuine operational shutdowns; flagged\n",
    "for separate sensitivity analysis rather than deletion.\n")
```

```{r eda-skewkurt-r}
sk_fn <- function(x) {
  x <- x[!is.na(x)]; n <- length(x); m <- mean(x); s <- sd(x)
  sum((x - m)^3) / (n * s^3)
}
kt_fn <- function(x) {
  x <- x[!is.na(x)]; n <- length(x); m <- mean(x); s <- sd(x)
  sum((x - m)^4) / (n * s^4) - 3
}
sk <- sk_fn(prod_r$shortfall)
kt <- kt_fn(prod_r$shortfall)
cat(sprintf("Production shortfall — Skewness: %.3f | Excess Kurtosis: %.3f\n", sk, kt))
cat(ifelse(sk < -0.5,
  "Left-skewed: underperformance days are more extreme and frequent than overperformance.\n",
  ifelse(sk > 0.5, "Right-skewed distribution.\n", "Approximately symmetric distribution.\n")))
cat(ifelse(kt > 1,
  "Leptokurtic (heavy tails): extreme shortfall days more frequent than normal,\n consistent with maintenance events and force majeure.\n",
  "Near-normal or platykurtic tails.\n"))
```

```{r eda-summary-r}
prod_r |>
  group_by(ZONE, FLOW_STATIONS) |>
  summarise(
    n              = n(),
    mean_budget    = mean(BUDGET,    na.rm = TRUE),
    mean_actual    = mean(ACTUAL,    na.rm = TRUE),
    mean_shortfall = mean(shortfall, na.rm = TRUE),
    pct_days_under = mean(shortfall < 0) * 100,
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric), ~ round(., 1))) |>
  kable(caption = "EDA Summary: mean daily production and shortfall by zone and station",
        col.names = c("Zone", "Station", "n", "Avg Budget",
                      "Avg Actual", "Avg Shortfall", "% Days Under Budget")) |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Python

```{python eda-anscombe-py}
# Anscombe's Quartet in Python to reinforce visual-diagnostic necessity
ans_py = pd.DataFrame({
  "x1": [10,8,13,9,11,14,6,4,12,7,5],
  "y1": [8.04,6.95,7.58,8.81,8.33,9.96,7.24,4.26,10.84,4.82,5.68],
  "x2": [10,8,13,9,11,14,6,4,12,7,5],
  "y2": [9.14,8.14,8.74,8.77,9.26,8.10,6.13,3.10,9.13,7.26,4.74],
  "x3": [10,8,13,9,11,14,6,4,12,7,5],
  "y3": [7.46,6.77,12.74,7.11,7.81,8.84,6.08,5.39,8.15,6.42,5.73],
  "x4": [8,8,8,8,8,8,8,19,8,8,8],
  "y4": [6.58,5.76,7.71,8.84,8.47,7.04,5.25,12.50,5.56,7.91,6.89]
})

rows = []
for i in [1, 2, 3, 4]:
  rows.append({
    "set": f"Set {i}",
    "mean_x": ans_py[f"x{i}"].mean(),
    "mean_y": ans_py[f"y{i}"].mean(),
    "corr_xy": ans_py[[f"x{i}", f"y{i}"]].corr().iloc[0, 1],
  })
display(pd.DataFrame(rows).round(3))

fig, axes = plt.subplots(2, 2, figsize=(8, 7))
for i, ax in enumerate(axes.flatten(), start=1):
  x = ans_py[f"x{i}"]
  y = ans_py[f"y{i}"]
  ax.scatter(x, y, s=25, alpha=0.85)
  m, b = np.polyfit(x, y, 1)
  xx = np.linspace(x.min(), x.max(), 50)
  ax.plot(xx, m * xx + b, color="red", linewidth=1)
  ax.set_title(f"Set {i}")
  ax.set_xlabel("x")
  ax.set_ylabel("y")
plt.suptitle("Anscombe's Quartet: identical summaries, different structures", y=1.02)
plt.tight_layout()
plt.show()
```

```{python eda-py}
zones_list = prod_py["ZONE"].dropna().unique()
fig, axes = plt.subplots(1, len(zones_list), figsize=(6 * len(zones_list), 4), squeeze=False)

for ax, zone in zip(axes[0], zones_list):
    subset = prod_py[prod_py["ZONE"] == zone]["shortfall"]
    ax.hist(subset, bins=40, alpha=0.7, edgecolor="white")
    ax.axvline(0, color="red", linestyle="--", label="On-target")
    ax.set_title(f"Shortfall Distribution — {zone}")
    ax.set_xlabel("Shortfall (bbl/day)")
    ax.set_ylabel("Count")
    ax.legend()

plt.suptitle("Daily Production Shortfall by Zone", fontsize=12, fontweight="bold")
plt.tight_layout()
plt.show()
```

```{python eda-outliers-py}
from scipy.stats import iqr as scipy_iqr

outlier_rows = []
for station, grp in prod_py.groupby("FLOW_STATIONS"):
    q1 = grp["shortfall"].quantile(0.25)
    q3 = grp["shortfall"].quantile(0.75)
    iqr_val = q3 - q1
    n_out = ((grp["shortfall"] < q1 - 1.5 * iqr_val) |
             (grp["shortfall"] > q3 + 1.5 * iqr_val)).sum()
    outlier_rows.append({"Station": station, "Q1": round(q1), "Q3": round(q3),
                         "IQR": round(iqr_val), "Lower_Fence": round(q1 - 1.5*iqr_val),
                         "Upper_Fence": round(q3 + 1.5*iqr_val),
                         "n_outliers": n_out,
                         "pct_outliers": round(n_out / len(grp) * 100, 1)})

print("IQR-based outlier detection by flow station:")
display(pd.DataFrame(outlier_rows))
print("\nDecision: retain extreme observations as genuine shutdowns/force-majeure events.")
```

```{python eda-skewkurt-py}
from scipy.stats import skew, kurtosis as sp_kurtosis

sk_val = skew(prod_py["shortfall"].dropna())
kt_val = sp_kurtosis(prod_py["shortfall"].dropna())
print(f"Production shortfall — Skewness: {sk_val:.3f} | Excess Kurtosis: {kt_val:.3f}")
print("Left-skewed: underperformance days dominate." if sk_val < -0.5 else
      ("Right-skewed." if sk_val > 0.5 else "Approximately symmetric."))
print("Leptokurtic (heavy tails — extreme shortfalls more frequent than normal)." if kt_val > 1
      else "Near-normal or platykurtic tails.")
```

```{python eda-summary-py}
summary = prod_py.groupby(["ZONE", "FLOW_STATIONS"]).agg(
    n=("shortfall", "count"),
    mean_budget=("BUDGET", "mean"),
    mean_actual=("ACTUAL", "mean"),
    mean_shortfall=("shortfall", "mean"),
    pct_days_under=("shortfall", lambda x: (x < 0).mean() * 100)
).round(1).reset_index()
display(summary)
```

:::

**Interpretation:** The shortfall distribution is left-skewed, meaning underperformance days are both more frequent and more extreme than overperformance days. Across the full sample, 86.4% of station-days fall below budget — confirming systemic production pressure rather than isolated volatility. IQR-based outlier detection flags a meaningful share of records in each station as extreme low values; these are retained in all analyses as genuine operational shutdowns rather than data errors, but are noted as potential influence points in regression. The Anscombe demonstration above reinforces why this visual step is mandatory: identical summary moments can conceal very different data shapes. At station level, CC2 has the largest mean deficit (−2,837.63 bbl/day), while CC3 has the highest share of under-budget days (92.8%), indicating highly persistent reliability constraints. Zone-level averages show sustained deficits (Zone 1: −2,198 bbl/day; Zone 2: −1,830 bbl/day). This pattern indicates chronic throughput pressure concentrated in specific assets; corrective action should prioritise persistent bottlenecks (CC2 and CC3) rather than one-off incident response.

---

## Technique 2 — Data Visualization

**Theory recap:** Tufte's (1983) principle of maximising the data-ink ratio and Wilkinson's (2005) Grammar of Graphics — implemented in R as `ggplot2` and in Python as `seaborn`/`matplotlib` — guide chart construction. Each chart type should be selected to match the data structure: time series for trends, heat maps for cross-tabulated magnitudes, scatter plots for bivariate relationships, box plots for distributional comparisons.

**Business justification:** Time-series and comparative charts communicate throughput trends and zone performance to management in the format used in board briefings. The five plots below form a single coherent narrative: systemic underperformance is worsening over time, is concentrated in Zone 1 stations, and coincides with periods when evacuation capacity appears constrained.

::: {.panel-tabset}

## R

```{r viz-timeseries-r}
p_ts <- ggplot(df_r, aes(x = DATE)) +
  geom_line(aes(y = ACTUAL_zone, colour = "Actual"),  linewidth = 0.8) +
  geom_line(aes(y = BUDGET_zone, colour = "Budget"), linetype = "dashed", linewidth = 0.8) +
  facet_wrap(~ZONE, ncol = 1, scales = "free_y") +
  scale_y_continuous(labels = comma) +
  scale_colour_manual(values = c("Actual" = "#2c7bb6", "Budget" = "#d7191c")) +
  labs(title    = "Daily Production: Actual vs Budget by Zone",
       subtitle = "Jan–May 2026 (May = partial, 5 days only)",
       x = "Date", y = "Production (bbl/day)", colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
p_ts
```

```{r viz-utilization-r}
p_util <- ggplot(evac_r, aes(x = DATE, y = utilization_rate, colour = ZONE)) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = 1/1.5, linetype = "dashed", colour = "orange",
             linewidth = 0.8) +
  annotate("text", x = min(evac_r$DATE) + 5, y = 1/1.5 + 0.02,
           label = "Max acceptable utilization (66.7% = 1/1.5 capacity buffer)",
           hjust = 0, size = 2.8) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Evacuation Utilization Rate by Zone",
       x = "Date", y = "Utilization (util / capacity)", colour = "Zone") +
  theme_minimal(base_size = 11)
p_util
```

```{r viz-heatmap-r}
p_heat <- prod_r |>
  mutate(month_label = format(DATE, "%b %Y")) |>
  group_by(FLOW_STATIONS, month_label) |>
  summarise(avg_shortfall = mean(shortfall, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = month_label, y = FLOW_STATIONS, fill = avg_shortfall)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = comma(round(avg_shortfall, 0))), size = 3) +
  scale_fill_gradient2(low = "#d73027", mid = "white", high = "#1a9850",
                       midpoint = 0, labels = comma, name = "Avg Shortfall") +
  labs(title = "Average Daily Shortfall Heatmap by Station and Month",
       caption = "May 2026 covers 5 days only — interpret with caution",
       x = "Month", y = "Flow Station") +
  theme_minimal(base_size = 11)
p_heat
```

```{r viz-scatter-r}
# Scatter: evacuation capacity vs zone-level shortfall, with fitted line
p_scatter <- ggplot(df_r |> drop_na(evacuation_capacity, shortfall_zone),
                    aes(x = evacuation_capacity, y = shortfall_zone, colour = ZONE)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(title = "Evacuation Capacity vs Production Shortfall",
       subtitle = "Negative slope confirms: higher capacity → smaller deficit",
       x = "Evacuation Capacity (bbl/day)", y = "Zone Shortfall (bbl/day)",
       colour = "Zone") +
  theme_minimal(base_size = 11)
p_scatter
```

```{r viz-monthly-r}
# Monthly mean shortfall trend — explicit trend line to highlight deterioration
p_monthly <- prod_r |>
  filter(month(DATE) < 5 | (month(DATE) == 5 & day(DATE) <= 5)) |>
  mutate(month_num  = floor_date(DATE, "month"),
         month_label = format(month_num, "%b")) |>
  group_by(ZONE, month_label, month_num) |>
  summarise(mean_shortfall = mean(shortfall, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = month_num, y = mean_shortfall, colour = ZONE, group = ZONE)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_x_datetime(date_labels = "%b", date_breaks = "1 month") +
  scale_y_continuous(labels = comma) +
  annotate("text", x = as.POSIXct("2026-05-01"), y = -500,
           label = "May = 5 days\nonly", size = 2.8, hjust = 0.5, colour = "grey40") +
  labs(title = "Monthly Mean Shortfall Trend by Zone (Jan–Apr complete; May partial)",
       x = "Month", y = "Mean Daily Shortfall (bbl/day)", colour = "Zone") +
  theme_minimal(base_size = 11)
p_monthly
```

```{r viz-combined-r}
# ── Cohesive five-panel patchwork layout ──────────────────────────────────────
(p_ts | p_util) /
(p_heat) /
(p_scatter | p_monthly) +
  plot_annotation(
    title   = "OML 18 Production Performance — Visual Narrative",
    subtitle = "Five-panel view: trend → capacity → station-month heat → bivariate driver → deterioration",
    theme   = theme(plot.title    = element_text(size = 13, face = "bold"),
                    plot.subtitle = element_text(size = 9, colour = "grey50"))
  )
```

```{r viz-trendtest-r}
# Formal trend check on complete months only (Jan-Apr)
trend_df <- prod_r |>
  filter(month(DATE) <= 4) |>
  mutate(month_idx = as.integer(factor(floor_date(DATE, "month"), ordered = TRUE))) |>
  group_by(month_idx) |>
  summarise(mean_shortfall = mean(shortfall, na.rm = TRUE), .groups = "drop")

trend_model <- lm(mean_shortfall ~ month_idx, data = trend_df)
trend_sum <- summary(trend_model)
cat("=== Monthly Shortfall Trend Test (Jan-Apr only) ===\n")
cat(sprintf("Slope: %.2f bbl/day per month | p-value: %.4f | R2: %.3f\n",
            coef(trend_model)["month_idx"],
            coef(trend_sum)["month_idx", "Pr(>|t|)"],
            trend_sum$r.squared))
```

## Python

```{python viz-timeseries-py}
fig, axes = plt.subplots(2, 1, figsize=(12, 7), sharex=True)
for ax, zone in zip(axes, df_py["ZONE"].unique()):
    z = df_py[df_py["ZONE"] == zone].sort_values("DATE")
    ax.plot(z["DATE"], z["ACTUAL_zone"], label="Actual", color="#2c7bb6", linewidth=1.2)
    ax.plot(z["DATE"], z["BUDGET_zone"], label="Budget", color="#d7191c",
            linestyle="--", linewidth=1.2)
    ax.set_title(f"Actual vs Budget — {zone}")
    ax.yaxis.set_major_formatter(mticker.FuncFormatter(lambda x, _: f"{x:,.0f}"))
    ax.legend()
plt.xlabel("Date")
fig.suptitle("Daily Production: Actual vs Budget by Zone (Jan–May 2026; May = 5 days partial)",
             fontsize=12, fontweight="bold")
plt.tight_layout()
plt.show()
```

```{python viz-utilization-py}
fig, ax = plt.subplots(figsize=(12, 4))
for zone, grp in evac_py.groupby("ZONE"):
    grp = grp.sort_values("DATE")
    ax.plot(grp["DATE"], grp["utilization_rate"], label=zone, linewidth=1.2)
ax.axhline(1/1.5, color="orange", linestyle="--", linewidth=1.2,
           label="Max acceptable (66.7% utilization)")
ax.yaxis.set_major_formatter(mticker.PercentFormatter(xmax=1))
ax.set_title("Evacuation Utilization Rate by Zone")
ax.set_xlabel("Date")
ax.set_ylabel("Utilization Rate")
ax.legend()
plt.tight_layout()
plt.show()
```

```{python viz-heatmap-py}
heat = prod_py.copy()
heat["month_label"] = heat["DATE"].dt.to_period("M").astype(str)
pivot = heat.groupby(["FLOW_STATIONS", "month_label"])["shortfall"].mean().unstack()

fig, ax = plt.subplots(figsize=(9, 4))
sns.heatmap(pivot, cmap="RdYlGn", center=0, annot=True, fmt=".0f",
            linewidths=0.5, ax=ax, cbar_kws={"label": "Avg Shortfall (bbl/day)"})
ax.set_title("Average Daily Shortfall Heatmap by Station and Month\n(May = 5 days only)")
plt.tight_layout()
plt.show()
```

```{python viz-scatter-py}
df_sc = df_py.dropna(subset=["evacuation_capacity", "shortfall_zone"])
fig, ax = plt.subplots(figsize=(8, 5))
for zone, grp in df_sc.groupby("ZONE"):
    ax.scatter(grp["evacuation_capacity"], grp["shortfall_zone"],
               label=zone, alpha=0.4, s=20)
    # Fitted line
    m, b = np.polyfit(grp["evacuation_capacity"], grp["shortfall_zone"], 1)
    x_range = np.linspace(grp["evacuation_capacity"].min(),
                          grp["evacuation_capacity"].max(), 100)
    ax.plot(x_range, m * x_range + b, linewidth=1.5)

ax.axhline(0, linestyle="--", color="grey")
ax.set_title("Evacuation Capacity vs Zone Shortfall\n(Negative slope = higher capacity → smaller deficit)")
ax.set_xlabel("Evacuation Capacity (bbl/day)")
ax.set_ylabel("Zone Shortfall (bbl/day)")
ax.legend()
ax.xaxis.set_major_formatter(mticker.FuncFormatter(lambda x, _: f"{x:,.0f}"))
ax.yaxis.set_major_formatter(mticker.FuncFormatter(lambda x, _: f"{x:,.0f}"))
plt.tight_layout()
plt.show()
```

```{python viz-monthly-py}
monthly = prod_py.copy()
monthly = monthly[~((monthly["DATE"].dt.month == 5) & (monthly["DATE"].dt.day > 5))]
monthly["month"] = monthly["DATE"].dt.to_period("M").dt.to_timestamp()
monthly_grp = monthly.groupby(["ZONE", "month"])["shortfall"].mean().reset_index()

fig, ax = plt.subplots(figsize=(9, 4))
for zone, grp in monthly_grp.groupby("ZONE"):
    ax.plot(grp["month"], grp["shortfall"], marker="o", label=zone, linewidth=1.3)
ax.axhline(0, linestyle="--", color="grey")
ax.annotate("May = 5 days", xy=(monthly_grp["month"].max(), -500),
            fontsize=8, color="grey")
ax.set_title("Monthly Mean Shortfall Trend — Jan–Apr complete; May partial")
ax.set_xlabel("Month")
ax.set_ylabel("Mean Daily Shortfall (bbl/day)")
ax.yaxis.set_major_formatter(mticker.FuncFormatter(lambda x, _: f"{x:,.0f}"))
ax.legend()
plt.tight_layout()
plt.show()
```

```{python viz-trendtest-py}
# Formal trend check on complete months only (Jan-Apr)
trend = prod_py[prod_py["DATE"].dt.month <= 4].copy()
trend["month_idx"] = trend["DATE"].dt.month - trend["DATE"].dt.month.min() + 1
trend_grp = trend.groupby("month_idx", as_index=False)["shortfall"].mean()

trend_mod = smf.ols("shortfall ~ month_idx", data=trend_grp).fit()
print("=== Monthly Shortfall Trend Test (Jan-Apr only) ===")
print(f"Slope: {trend_mod.params['month_idx']:.2f} bbl/day per month")
print(f"p-value: {trend_mod.pvalues['month_idx']:.4f} | R-squared: {trend_mod.rsquared:.3f}")
```

:::

**Visualization narrative (five plots, one story):**

1. **Time series (Actual vs Budget):** Both zones run persistently below budget from January onwards, with the gap widening over time — no evidence of seasonal recovery.
2. **Utilization rate:** Both zones operate comfortably below the 66.7% maximum acceptable threshold on most days, demonstrating that evacuation capacity *headroom exists* but is not translating into throughput recovery. The primary constraint is therefore upstream of the evacuation route.
3. **Monthly shortfall heat map:** CC2 and CC3 show the deepest red cells across all months; CC1 and ALK are relatively less severe. Worsening is consistent across all four stations from January to April. May cells should be treated with caution (5 days only).
4. **Scatter plot (capacity vs shortfall):** The negative fitted lines confirm that higher evacuation capacity is associated with smaller deficits across both zones — the key bivariate relationship that regression will subsequently quantify and control.
5. **Monthly trend line:** The mean shortfall deepens progressively each month in both zones, and the Jan-Apr formal trend test confirms the deterioration is statistically directional rather than purely visual. This supports the interpretation of a structural performance decline consistent with ageing well productivity or deferred maintenance.

---

## Technique 3 — Hypothesis Testing

**Theory recap:** The one-way ANOVA tests whether group means differ more than chance variation would predict (Fisher, 1921). Its F-statistic is the ratio of between-group variance to within-group variance. Assumptions include normality within groups and homogeneity of variance. When normality is violated (as confirmed by Shapiro-Wilk), the Kruskal-Wallis H-test — a non-parametric rank-based equivalent — is the appropriate primary reference. Effect size (η²) contextualises statistical significance: a small p-value in a large sample can reflect a trivially small real difference.

**Business justification:** Formally tests whether Zone 1 and Zone 2 perform statistically differently, converting an operational observation into a defensible basis for resource allocation decisions.

---

**Hypothesis 1 — Zone-level shortfall difference**

> **H₀:** Mean daily production shortfall is equal across evacuation zones (μ_Zone1 = μ_Zone2).
> **H₁:** Mean daily production shortfall differs significantly between evacuation zones.
> **Significance level:** α = 0.05.

---

**Hypothesis 2 — Station-level shortfall difference**

> **H₀:** Mean daily production shortfall is equal across all four flow stations (μ_ALK = μ_CC1 = μ_CC2 = μ_CC3).
> **H₁:** At least one flow station has a significantly different mean shortfall.
> **Significance level:** α = 0.05.

---

::: {.panel-tabset}

## R

```{r hyp-r}
# ── Assumption check: normality ───────────────────────────────────────────────
zones_r <- sort(unique(prod_r$ZONE))
cat("=== Shapiro-Wilk Normality Test (required before choosing parametric vs non-parametric) ===\n")
for (z in zones_r) {
  sw <- shapiro.test(prod_r$shortfall[prod_r$ZONE == z])
  cat(sprintf("  %s: W=%.4f, p=%.4f %s\n", z, sw$statistic, sw$p.value,
              ifelse(sw$p.value < 0.05, "→ NON-NORMAL (use non-parametric)", "→ Normal")))
}

# ── Test 1: ANOVA (parametric) ────────────────────────────────────────────────
cat("\n=== ANOVA: Shortfall by Zone ===\n")
aov_zone <- aov(shortfall ~ ZONE, data = prod_r)
aov_tbl  <- summary(aov_zone)
print(aov_tbl[[1]])
eta_sq <- aov_tbl[[1]]["ZONE", "Sum Sq"] / sum(aov_tbl[[1]][, "Sum Sq"])
cat(sprintf("\nEta-squared (effect size): %.4f → %s effect\n", eta_sq,
            ifelse(eta_sq < 0.01, "negligible",
            ifelse(eta_sq < 0.06, "small",
            ifelse(eta_sq < 0.14, "medium", "large")))))

# ── Test 1 robustness: Kruskal-Wallis (non-parametric, preferred given non-normality) ──
cat("\n=== Kruskal-Wallis (primary reference — normality violated) ===\n")
kw <- kruskal.test(shortfall ~ ZONE, data = prod_r)
cat(sprintf("H = %.4f, df = %d, p = %.4f → %s\n", kw$statistic, kw$parameter, kw$p.value,
            ifelse(kw$p.value < 0.05, "Reject H₀", "Fail to reject H₀")))
```

```{r hyp-station-r}
# ── Test 2: Station-level ANOVA ───────────────────────────────────────────────
cat("=== ANOVA: Shortfall by Flow Station ===\n")
aov_station <- aov(shortfall ~ FLOW_STATIONS, data = prod_r)
aov_stn_tbl <- summary(aov_station)
print(aov_stn_tbl[[1]])
eta_sq_stn <- aov_stn_tbl[[1]]["FLOW_STATIONS", "Sum Sq"] / sum(aov_stn_tbl[[1]][, "Sum Sq"])
cat(sprintf("\nEta-squared (station level): %.4f → %s effect\n", eta_sq_stn,
            ifelse(eta_sq_stn < 0.01, "negligible",
            ifelse(eta_sq_stn < 0.06, "small",
            ifelse(eta_sq_stn < 0.14, "medium", "large")))))

# Tukey HSD post-hoc pairwise comparisons
cat("\n=== Tukey HSD — pairwise station comparisons ===\n")
print(TukeyHSD(aov_station))
```

## Python

```{python hyp-py}
from scipy.stats import f_oneway, kruskal, shapiro

zones = prod_py["ZONE"].unique()
groups = [prod_py[prod_py["ZONE"] == z]["shortfall"].dropna().values for z in zones]

print("=== Shapiro-Wilk Normality Test ===")
for z, g in zip(zones, groups):
    stat, p = shapiro(g[:5000])
    print(f"  {z}: W={stat:.4f}, p={p:.4f} {'→ NON-NORMAL' if p < 0.05 else '→ Normal'}")

f_stat, p_anova = f_oneway(*groups)
print(f"\n=== One-way ANOVA: Shortfall by Zone ===")
print(f"F = {f_stat:.4f}, p = {p_anova:.4f} → {'Reject H₀' if p_anova < 0.05 else 'Fail to reject H₀'}")

h_stat, p_kw = kruskal(*groups)
print(f"\n=== Kruskal-Wallis (primary, non-parametric) ===")
print(f"H = {h_stat:.4f}, p = {p_kw:.4f} → {'Reject H₀' if p_kw < 0.05 else 'Fail to reject H₀'}")
```

```{python hyp-station-py}
from scipy.stats import f_oneway
from statsmodels.stats.multicomp import pairwise_tukeyhsd

stations = prod_py["FLOW_STATIONS"].unique()
station_groups = [prod_py[prod_py["FLOW_STATIONS"] == s]["shortfall"].dropna().values
                  for s in stations]

f2, p2 = f_oneway(*station_groups)
print(f"=== ANOVA: Shortfall by Flow Station ===")
print(f"F = {f2:.4f}, p = {p2:.4f} → {'Reject H₀' if p2 < 0.05 else 'Fail to reject H₀'}")

# Eta-squared
ss_between = sum(len(g) * (g.mean() - np.concatenate(station_groups).mean())**2 for g in station_groups)
ss_total   = sum(((x - np.concatenate(station_groups).mean())**2).sum() for x in station_groups)
eta_sq_stn = ss_between / ss_total
print(f"Eta-squared (station level): {eta_sq_stn:.4f} → {'medium' if eta_sq_stn >= 0.06 else 'small'} effect")

print("\n=== Tukey HSD pairwise ===")
tukey_result = pairwise_tukeyhsd(
    endog=prod_py["shortfall"].dropna(),
    groups=prod_py.loc[prod_py["shortfall"].notna(), "FLOW_STATIONS"]
)
print(tukey_result.summary())
```

:::

**Interpretation and decisions:**

| Test | Statistic | p-value | Decision | Effect Size |
|------|-----------|---------|----------|-------------|
| ANOVA — Zone | F = 4.083 | 0.044 | Reject H₀ at α = 0.05 | η² = 0.007 (negligible) |
| Kruskal-Wallis — Zone | H = 1.678 | 0.195 | **Fail to reject H₀** | — |
| ANOVA — Station | F = 32.599 | < 0.001 | Reject H₀ | η² ≈ 0.14 (medium–large) |

The Shapiro-Wilk test confirms non-normality in both zones (p < 0.05), which violates the key ANOVA assumption. The Kruskal-Wallis test — which requires no normality assumption and is therefore the appropriate primary reference — returns p = 0.195: a non-significant result. The zone-level difference should be treated as **directionally suggestive but not statistically robust** once distributional properties are properly accounted for. The ANOVA eta-squared of 0.007 reinforces this: zone membership explains less than 1% of total shortfall variance.

At station level, the ANOVA is strongly significant and robust given larger group sizes; eta-squared (~0.14) indicates a medium-to-large effect. Tukey HSD post-hoc comparisons identify CC2 as statistically distinct from ALK and CC1. **Operational implication: station-level targeting — specifically CC2 and CC3 within Zone 1 — is the primary evidence-based justification for resource allocation, not zone-level comparisons alone.**

To reconcile the apparent zone contradiction: the ANOVA zone p-value (0.044) is sensitive to violated normality, while the Kruskal-Wallis result (0.195) is robust under the observed distribution shape and is therefore treated as primary for zone-level inference. Regression zone coefficients are interpreted conditionally (holding capacity/utilization constant), not as standalone proof of unconditional zone differences.

---

## Technique 4 — Correlation Analysis

**Theory recap:** Pearson's r measures the strength and direction of a *linear* relationship between two continuous variables (range: −1 to +1). Spearman's ρ is the rank-based equivalent — appropriate when normality is violated (as confirmed here) or when relationships may be monotonic but non-linear. Kendall's τ is a slower-converging but more robust rank correlation. Partial correlation removes the shared influence of a third variable to estimate the unique association between two variables of interest (Pearson, 1900).

**Business justification:** Maps relationships between budget targets, actual production, evacuation capacity, and utilization before building the regression model. Given confirmed non-normality, Spearman is the primary measure; Pearson is reported for comparability. Partial correlation isolates the unique relationship between `evacuation_capacity` and `shortfall` after removing the confounding influence of zone-level budget scale.

::: {.panel-tabset}

## R

```{r corr-r}
corr_vars <- df_r |>
  select(BUDGET_zone, ACTUAL_zone, shortfall_zone,
         evacuation_capacity, evacuation_utilization, utilization_rate) |>
  drop_na()

# Pearson, Spearman, and Kendall matrices
cor_pearson  <- cor(corr_vars, method = "pearson")
cor_spearman <- cor(corr_vars, method = "spearman")
cor_kendall  <- cor(corr_vars, method = "kendall")

# ── Tidy heatmap using ggplot2 geom_tile (avoids stringi dependency) ──────────
gg_corr_tile(cor_pearson,  "Pearson Correlation Matrix — Production & Evacuation Variables")
```

```{r corr-spearman-r}
gg_corr_tile(cor_spearman, "Spearman Rank Correlation Matrix (preferred: non-normal data)")
```

```{r corr-kendall-r}
gg_corr_tile(cor_kendall, "Kendall Tau Correlation Matrix (robust rank concordance)")
```

```{r corr-table-r}
# Ranked correlation table with both methods
cor_compare <- data.frame(
  Variable = rownames(cor_pearson),
  Pearson_r   = cor_pearson[, "shortfall_zone"],
  Spearman_rho = cor_spearman[, "shortfall_zone"],
  Kendall_tau  = cor_kendall[, "shortfall_zone"]
) |>
  filter(Variable != "shortfall_zone") |>
  arrange(desc(abs(Spearman_rho)))

kable(cor_compare |> mutate(across(where(is.numeric), ~ round(., 4))),
      caption = "Pearson vs Spearman vs Kendall correlation with zone-level daily shortfall") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r partial-corr-r}
# ── Partial correlation: evacuation_capacity ~ shortfall controlling for BUDGET_zone ──
# Method: regress each variable on the control, use residuals
model_e <- lm(evacuation_capacity ~ BUDGET_zone, data = corr_vars)
model_s <- lm(shortfall_zone      ~ BUDGET_zone, data = corr_vars)
partial_r <- cor(residuals(model_e), residuals(model_s))
cat(sprintf("Partial correlation (evacuation_capacity ~ shortfall | BUDGET_zone removed): r = %.4f\n",
            partial_r))
cat("Interpretation: After removing the budget-scale effect, evacuation capacity retains a\n",
    "meaningful independent association with shortfall.\n")
```

## Python

```{python corr-py}
corr_cols = ["BUDGET_zone", "ACTUAL_zone", "shortfall_zone",
             "evacuation_capacity", "evacuation_utilization", "utilization_rate"]
corr_data = df_py[corr_cols].dropna()

cor_pearson_py  = corr_data.corr(method="pearson")
cor_spearman_py = corr_data.corr(method="spearman")
cor_kendall_py  = corr_data.corr(method="kendall")

fig, axes = plt.subplots(1, 3, figsize=(22, 6))

for ax, mat, title in zip(axes,
                           [cor_pearson_py, cor_spearman_py, cor_kendall_py],
                           ["Pearson Correlation", "Spearman Rank Correlation (preferred)", "Kendall Tau Correlation"]):
    mask = np.triu(np.ones_like(mat, dtype=bool))
    sns.heatmap(mat, mask=mask, annot=True, fmt=".2f", cmap="RdYlGn",
                center=0, vmin=-1, vmax=1, linewidths=0.5, ax=ax,
                cbar_kws={"shrink": 0.8})
    ax.set_title(title, fontsize=11, fontweight="bold")

plt.tight_layout()
plt.show()
```

```{python partial-corr-py}
import statsmodels.formula.api as smf_partial

# Partial correlation: remove BUDGET_zone influence from both variables
res_cap  = smf.ols("evacuation_capacity ~ BUDGET_zone", data=corr_data).fit().resid
res_shrt = smf.ols("shortfall_zone      ~ BUDGET_zone", data=corr_data).fit().resid

partial_r_py, p_partial = stats.pearsonr(res_cap, res_shrt)
print(f"Partial correlation (evacuation_capacity ~ shortfall | BUDGET_zone controlled): r = {partial_r_py:.4f}, p = {p_partial:.4f}")
print("After removing budget-scale effect, the independent capacity–shortfall relationship is retained.")

# Comparison table
print("\nCorrelation with shortfall_zone (Pearson vs Spearman vs Kendall):")
compare_df = pd.DataFrame({
    "Variable": [v for v in corr_cols if v != "shortfall_zone"],
    "Pearson_r": [cor_pearson_py.loc[v, "shortfall_zone"] for v in corr_cols if v != "shortfall_zone"],
  "Spearman_rho": [cor_spearman_py.loc[v, "shortfall_zone"] for v in corr_cols if v != "shortfall_zone"],
  "Kendall_tau": [cor_kendall_py.loc[v, "shortfall_zone"] for v in corr_cols if v != "shortfall_zone"]
}).sort_values("Spearman_rho", key=abs, ascending=False)
display(compare_df.round(4))
```

:::

**Interpretation of the three strongest correlations with `shortfall_zone`:**

1. **`BUDGET_zone` (Pearson r ≈ −0.60):** Largest in magnitude, but this is partly a mathematical artifact. Since `shortfall = ACTUAL − BUDGET`, higher budget targets mechanically produce more negative shortfall values in absolute terms. This figure cannot be read as an independent operational lever. The partial correlation analysis removes this scale effect.

2. **`evacuation_capacity` (Pearson r ≈ −0.36; Spearman ρ ≈ −0.33; Kendall τ < 0):** Zones with higher absolute evacuation capacity consistently record smaller production deficits. Pearson, Spearman, and Kendall all agree in direction and practical magnitude, lending robustness to the finding under both parametric and rank-based assumptions. Partial correlation confirms the relationship survives after controlling for budget scale. This is the key actionable finding: increasing evacuation capacity is associated with improved production attainment.

3. **`utilization_rate` (Pearson r ≈ +0.18; Spearman ρ ≈ +0.15):** Only a weak positive association. The rate at which available capacity is used appears to matter less than the absolute capacity level itself — a finding the regression model confirms. This is operationally interpretable: having capacity headroom (high absolute capacity, moderate utilization) matters more than running the pipeline hard.

**Note on Pearson, Spearman, and Kendall:** Given confirmed non-normality, rank-based measures are preferred as primary references. Spearman and Kendall both align with Pearson on sign and relative strength, increasing confidence that the capacity-shortfall relationship is not an artifact of one metric choice.

---

## Technique 5 — Multiple Regression

**Theory recap:** Ordinary Least Squares (OLS) regression estimates the linear combination of predictors that minimises the sum of squared residuals. Key assumptions: linearity, independence of errors, homoscedasticity, approximate normality of residuals, and no severe multicollinearity. Violation of these assumptions affects the reliability of coefficient estimates and their standard errors. Diagnostics include residual plots (linearity, homoscedasticity), Q-Q plots (normality), Cook's distance (influential observations), Variance Inflation Factors (multicollinearity), and the Durbin-Watson statistic (autocorrelation).

**Business justification:** Quantifies the independent contribution of zone, evacuation capacity, and utilization to daily production shortfall. The resulting coefficients support concrete fleet-sizing and capacity-allocation recommendations to management.

::: {.panel-tabset}

## R

```{r reg-r}
df_r$ZONE_f <- factor(df_r$ZONE)

model <- lm(shortfall_zone ~ ZONE_f + evacuation_capacity + utilization_rate,
            data = df_r)

cat("=== OLS Regression: Daily Zone Shortfall ===\n")
summary(model)
```

```{r reg-diagnostics-r}
coef_data <- data.frame(
  term      = names(coef(model)),
  estimate  = coef(model),
  std.error = sqrt(diag(vcov(model))),
  row.names = NULL
)
coef_data$t.value  <- coef_data$estimate / coef_data$std.error
coef_data$p.value  <- 2 * pt(abs(coef_data$t.value), df = model$df.residual,
                               lower.tail = FALSE)
coef_data$conf.low  <- coef_data$estimate - 1.96 * coef_data$std.error
coef_data$conf.high <- coef_data$estimate + 1.96 * coef_data$std.error
coef_data[, -1] <- round(coef_data[, -1], 4)

kable(coef_data, caption = "Regression coefficients with 95% confidence intervals") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

model_summary <- summary(model)
fit_data <- data.frame(
  r.squared     = round(model_summary$r.squared, 4),
  adj.r.squared = round(model_summary$adj.r.squared, 4),
  fstatistic    = round(model_summary$fstatistic[1], 4),
  p.value       = round(pf(model_summary$fstatistic[1], model_summary$fstatistic[2],
                            model_summary$fstatistic[3], lower.tail = FALSE), 6),
  AIC = round(AIC(model), 2),
  BIC = round(BIC(model), 2)
)
kable(fit_data, caption = "Model fit statistics") |>
  kable_styling(full_width = FALSE)
```

```{r reg-plots-r}
par(mfrow = c(2, 2))
plot(model, which = 1:4, pch = 16, cex = 0.6)
par(mfrow = c(1, 1))
```

```{r reg-assumptions-r}
# VIF — manual computation (no car package required)
X_mat <- model.matrix(model)[, -1]
vif_vals <- sapply(seq_len(ncol(X_mat)), function(i) {
  r2_i <- summary(lm(X_mat[, i] ~ X_mat[, -i]))$r.squared
  1 / (1 - r2_i)
})
names(vif_vals) <- colnames(X_mat)
cat("=== Variance Inflation Factors (VIF) ===\n")
print(round(vif_vals, 3))
cat("Note: VIF > 5 = problematic multicollinearity; > 10 = severe\n\n")

# Durbin-Watson
resids <- residuals(model)
dw_stat <- sum(diff(resids)^2) / sum(resids^2)
cat(sprintf("=== Durbin-Watson Statistic: %.4f ===\n", dw_stat))
cat(ifelse(dw_stat < 1.5,
  "CAUTION: Positive autocorrelation — OLS SEs may be underestimated; consider robust SEs.\n",
  ifelse(dw_stat > 2.5, "CAUTION: Negative autocorrelation.\n",
  "No strong autocorrelation at this threshold.\n")))

# Newey-West HAC robust standard errors (acts on the autocorrelation warning)
hac_vcov <- sandwich::NeweyWest(model, lag = 5, prewhite = FALSE, adjust = TRUE)
hac_test <- lmtest::coeftest(model, vcov. = hac_vcov)
cat("\n=== HAC (Newey-West) Robust Coefficients ===\n")
print(round(hac_test, 4))

cap_beta_hac <- unname(hac_test["evacuation_capacity", "Estimate"])
cap_se_hac <- unname(hac_test["evacuation_capacity", "Std. Error"])
cap_ci_hac <- cap_beta_hac + c(-1.96, 1.96) * cap_se_hac
```

```{r reg-validation-r}
# Out-of-sample validation: train on Jan-Mar, test on Apr-May
split_date <- as.Date("2026-03-31")
train_r <- df_r |> filter(DATE <= split_date)
test_r  <- df_r |> filter(DATE > split_date)

model_train_r <- lm(shortfall_zone ~ ZONE_f + evacuation_capacity + utilization_rate,
                    data = train_r)
pred_test_r <- predict(model_train_r, newdata = test_r)

rmse_r <- sqrt(mean((test_r$shortfall_zone - pred_test_r)^2, na.rm = TRUE))
mae_r  <- mean(abs(test_r$shortfall_zone - pred_test_r), na.rm = TRUE)

cat("=== Holdout Validation (Train Jan-Mar | Test Apr-May) ===\n")
cat(sprintf("RMSE: %.2f bbl/day | MAE: %.2f bbl/day | Test n: %d\n",
            rmse_r, mae_r, nrow(test_r)))
cat(sprintf("Capacity coefficient stability (full vs train): %.4f vs %.4f\n",
            coef(model)["evacuation_capacity"],
            coef(model_train_r)["evacuation_capacity"]))
```

```{r reg-scenario-r}
# Capacity intervention scenario ranges from HAC-adjusted uncertainty
scenario_df <- tibble(
  capacity_increase = c(5000, 10000, 15000)
) |>
  mutate(
    daily_recovery_est = -cap_beta_hac * capacity_increase,
    daily_recovery_low = -cap_ci_hac[2] * capacity_increase,
    daily_recovery_high = -cap_ci_hac[1] * capacity_increase,
    annual_recovery_est = daily_recovery_est * 365,
    annual_recovery_pct_18m = annual_recovery_est / 18000000 * 100
  )

kable(
  scenario_df |>
    mutate(across(where(is.numeric), ~ round(., 2))),
  caption = "Capacity expansion scenarios with HAC-based uncertainty bounds"
) |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Python

```{python reg-py}
df_reg = df_py.dropna(subset=["shortfall_zone", "evacuation_capacity",
                               "utilization_rate", "ZONE"])
df_reg = df_reg.copy()
df_reg["ZONE_f"] = pd.Categorical(df_reg["ZONE"])

model_py = smf.ols(
    "shortfall_zone ~ C(ZONE_f) + evacuation_capacity + utilization_rate",
    data=df_reg
).fit()
print(model_py.summary())
```

```{python reg-diagnostics-py}
fig, axes = plt.subplots(1, 2, figsize=(12, 4))

axes[0].scatter(model_py.fittedvalues, model_py.resid, alpha=0.5, s=15)
axes[0].axhline(0, color="red", linestyle="--")
axes[0].set_xlabel("Fitted values")
axes[0].set_ylabel("Residuals")
axes[0].set_title("Residuals vs Fitted")

sm.qqplot(model_py.resid, line="s", ax=axes[1], alpha=0.5)
axes[1].set_title("Q-Q Plot of Residuals")

plt.tight_layout()
plt.show()

print(f"R-squared: {model_py.rsquared:.4f}")
print(f"Adj. R-squared: {model_py.rsquared_adj:.4f}")
print(f"F-statistic p-value: {model_py.f_pvalue:.6f}")
```

```{python reg-assumptions-py}
from statsmodels.stats.outliers_influence import variance_inflation_factor
from statsmodels.stats.stattools import durbin_watson

X_vif = df_reg[["evacuation_capacity", "utilization_rate"]].dropna().reset_index(drop=True)
vif_df = pd.DataFrame({
    "Feature": X_vif.columns,
    "VIF": [round(variance_inflation_factor(X_vif.values, i), 3)
            for i in range(X_vif.shape[1])]
})
print("=== Variance Inflation Factors (VIF) ===")
print(vif_df.to_string(index=False))
print("Note: VIF > 5 = problematic; > 10 = severe\n")

dw_stat_py = durbin_watson(model_py.resid)
print(f"=== Durbin-Watson: {dw_stat_py:.4f} ===")
if dw_stat_py < 1.5:
    print("CAUTION: Positive autocorrelation — OLS SEs may be underestimated; consider HAC robust SEs.")
elif dw_stat_py > 2.5:
    print("CAUTION: Negative autocorrelation.")
else:
    print("No strong autocorrelation at this threshold.")

# HAC (Newey-West) robust covariance comparison
hac_py = model_py.get_robustcov_results(cov_type="HAC", maxlags=5)
hac_table = pd.DataFrame({
  "term": model_py.params.index,
  "coef_ols": model_py.params.values,
  "se_ols": model_py.bse.values,
  "se_hac": hac_py.bse,
  "p_ols": model_py.pvalues.values,
  "p_hac": hac_py.pvalues,
})
print("\n=== OLS vs HAC Robust SE Comparison ===")
print(hac_table.round(4).to_string(index=False))
```

```{python reg-validation-py}
# Out-of-sample validation: train on Jan-Mar, test on Apr-May
split_date = pd.Timestamp("2026-03-31")
train_py = df_reg[df_reg["DATE"] <= split_date].copy()
test_py = df_reg[df_reg["DATE"] > split_date].copy()

model_train_py = smf.ols(
  "shortfall_zone ~ C(ZONE_f) + evacuation_capacity + utilization_rate",
  data=train_py
).fit()

pred_test_py = model_train_py.predict(test_py)
rmse_py = np.sqrt(np.mean((test_py["shortfall_zone"] - pred_test_py) ** 2))
mae_py = np.mean(np.abs(test_py["shortfall_zone"] - pred_test_py))

print("=== Holdout Validation (Train Jan-Mar | Test Apr-May) ===")
print(f"RMSE: {rmse_py:.2f} bbl/day | MAE: {mae_py:.2f} bbl/day | Test n: {len(test_py)}")
print(f"Capacity coefficient stability (full vs train): {model_py.params['evacuation_capacity']:.4f} vs {model_train_py.params['evacuation_capacity']:.4f}")
```

```{python reg-scenario-py}
# Scenario analysis with HAC uncertainty bounds for evacuation capacity coefficient
cap_idx = list(model_py.params.index).index("evacuation_capacity")
cap_beta_hac_py = hac_py.params[cap_idx]
cap_se_hac_py = hac_py.bse[cap_idx]
cap_ci_hac_py = [cap_beta_hac_py - 1.96 * cap_se_hac_py, cap_beta_hac_py + 1.96 * cap_se_hac_py]

scenarios = pd.DataFrame({"capacity_increase": [5000, 10000, 15000]})
scenarios["daily_recovery_est"] = -cap_beta_hac_py * scenarios["capacity_increase"]
scenarios["daily_recovery_low"] = -cap_ci_hac_py[1] * scenarios["capacity_increase"]
scenarios["daily_recovery_high"] = -cap_ci_hac_py[0] * scenarios["capacity_increase"]
scenarios["annual_recovery_est"] = scenarios["daily_recovery_est"] * 365
scenarios["annual_recovery_pct_18m"] = scenarios["annual_recovery_est"] / 18000000 * 100

print("=== Capacity Expansion Scenarios (HAC-adjusted) ===")
display(scenarios.round(2))
```

:::

**Coefficient interpretation:**

| Term | Coefficient | Interpretation |
|------|-------------|---------------|
| Intercept | ~−2,000 | Baseline expected shortfall for Zone 1 at zero capacity/utilization |
| Zone 2 | +2,987.76*** | Conditional on equal capacity and utilization, Zone 2 shortfall is ~2,988 bbl/day *less severe* than Zone 1; this is a controlled-model contrast, not an unconditional zone test |
| `evacuation_capacity` | −0.5602*** | Each additional 1 bbl/day of capacity is associated with 0.56 bbl/day shortfall reduction; a 10,000 bbl/day increase → ~5,602 bbl/day recovery (~2.05 million barrels/year) |
| `utilization_rate` | Not significant (p = 0.755) | Once absolute capacity is controlled, the utilization rate adds no independent explanatory value |

**Model fit:** R² = 0.2904 (adj. = 0.2813); F-statistic p < 0.001. For operational daily data with many unmeasured factors (maintenance events, force majeure, staffing), 29% explained variance is acceptable for directional policy analysis, while 71% remains unexplained and is treated explicitly in the risk caveats.

**Diagnostic assessment:** The residuals vs fitted plot shows no strong systematic fan pattern (linearity and homoscedasticity broadly satisfied). The Q-Q plot shows mild tail deviation consistent with the confirmed non-normality; with n ≈ 250 zone-level observations this is within acceptable range under the central limit theorem. VIF values are below 5 (no severe multicollinearity). Durbin-Watson indicates likely positive autocorrelation, and HAC/Newey-West robust errors are therefore reported alongside OLS to guard inference quality. Holdout validation (train Jan-Mar, test Apr-May) provides out-of-sample RMSE/MAE checks to prevent in-sample overconfidence.

---

## Integrated Findings

The five analyses form a coherent, mutually reinforcing story about OML 18 crude oil production performance:

- **EDA** established that underperformance is persistent and widespread: 86.4% of station-days fall below budget, the shortfall distribution is left-skewed with heavy tails, and IQR-flagged extreme observations are consistent with maintenance shutdowns rather than data errors. CC2 shows the deepest average deficit; CC3 the highest persistence of underperformance.

- **Visualization** confirmed a worsening trajectory from January to May across all four stations, with mean monthly shortfalls deepening from −1,027 bbl/day (January) to −2,693 bbl/day (April). The scatter plot previewed the key finding: evacuation capacity has a meaningful negative slope against shortfall.

- **Hypothesis testing** refined the diagnostic picture. Zone-level differences, while significant under parametric ANOVA (F = 4.08, p = 0.044), are not supported by the more appropriate non-parametric Kruskal-Wallis test (H = 1.68, p = 0.195) and carry a negligible effect size (η² = 0.007). Station-level differences, by contrast, are robust and substantial (F = 32.6, p < 0.001; η² ≈ 0.14), with CC2 identified as statistically distinct from the better-performing stations.

- **Correlation analysis** confirmed — using both Pearson and Spearman to account for non-normality — that evacuation capacity is the most actionable operational variable correlated with shortfall (r ≈ −0.36). Partial correlation verified this association survives removal of the budget-scale confound. Utilization rate shows only a weak association, suggesting headroom matters more than throughput intensity.

- **Regression** quantified the operational lever: each 1 bbl/day of additional evacuation capacity is associated with a 0.56 bbl/day shortfall reduction, explaining 29% of total shortfall variance after controlling for zone and utilization. This coefficient remains directionally stable under HAC-robust errors and train/test validation checks.

**Single integrated recommendation:** Implement a targeted fleet expansion programme focused on CC2 and CC3 (within Zone 1) alternative evacuation routes. Under the observed data-generating conditions, a 10,000 bbl/day increase in effective evacuation capacity is associated with an expected recovery around 5,602 bbl/day; the scenario table reports HAC-based uncertainty bounds and 5k/10k/15k sensitivity cases. At the central estimate this is approximately 2.05 million additional barrels annually (about 11.4% of the 18 million barrel target). Because this is observational evidence, implementation should begin with a staged pilot and monitor realized response against the forecast band.

---

## Limitations & Further Work

1. **Partial month exposure (May 2026):** May data covers only 5 days. All monthly trend conclusions are expressed for January–April complete months; May figures are labelled throughout as partial and should be re-validated with full-month data before policy action.

2. **Non-linearity:** The regression assumes linear relationships. Threshold effects — where capacity below a critical floor triggers disproportionate shortfall — are plausible and should be explored with polynomial or spline terms. The scatter plots suggest some non-linearity in Zone 1 at low capacity values.

3. **Autocorrelation in daily data:** The Durbin-Watson statistic flags potential positive autocorrelation in the regression residuals, a common feature of daily time-series. HAC/Newey-West robust standard errors were therefore reported in Technique 5; although directional conclusions are stable, precision-sensitive decisions should prioritize the robust interval estimates.

4. **Three non-operational stations excluded:** ALK, CC1, CC2, CC3 are the four active stations. A full production performance model — particularly relevant if any of the three inactive stations are brought online — should incorporate those assets to avoid overstatement of Zone 1's structural shortfall.

5. **Unmeasured confounders:** The regression R² of 0.29 leaves 71% of shortfall variance unexplained. Key unobserved drivers likely include: scheduled and unscheduled maintenance events, well productivity decline curves, staffing levels, and force-majeure environmental events. Including maintenance event flags as binary predictors would substantially improve model fit.

6. **Model validation:** A Jan-Mar train / Apr-May test split was run to report out-of-sample RMSE and MAE. This provides an initial realism check, but a longer rolling-origin backtest is recommended before deploying the model for routine operational forecasting.

7. **Causal attribution:** Correlation and regression establish association, not causation. A natural experiment — for example, comparing production attainment in periods before and after a capacity expansion — would be required to establish a causal claim for the coefficient interpretation.

---

## Executive Dashboard (Management View)

This one-page dashboard consolidates the decision-critical outputs for executive review and rapid action planning.

::: {.panel-tabset}

## R

```{r dashboard-r}
dashboard_kpi <- tibble(
  KPI = c(
    "% Station-days below budget",
    "Zone test (Kruskal-Wallis p)",
    "Station test (ANOVA p)",
    "Capacity coefficient (bbl/day per 1 bbl/day)",
    "Model R-squared",
    "10k capacity scenario (daily)"
  ),
  Value = c(
    sprintf("%.1f%%", mean(prod_r$shortfall < 0, na.rm = TRUE) * 100),
    "0.195 (not significant)",
    "< 0.001 (significant)",
    sprintf("%.4f", coef(model)["evacuation_capacity"]),
    sprintf("%.3f", summary(model)$r.squared),
    "~5,602 bbl/day (with HAC range in scenario table)"
  )
)

kable(dashboard_kpi, caption = "Executive KPI dashboard") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Python

```{python dashboard-py}
kpi_df = pd.DataFrame({
    "KPI": [
        "% Station-days below budget",
        "Zone test (Kruskal-Wallis p)",
        "Station test (ANOVA p)",
        "Capacity coefficient (bbl/day per 1 bbl/day)",
        "Model R-squared",
        "10k capacity scenario (daily)",
    ],
    "Value": [
        f"{(prod_py['shortfall'] < 0).mean() * 100:.1f}%",
        "0.195 (not significant)",
        "< 0.001 (significant)",
        f"{model_py.params['evacuation_capacity']:.4f}",
        f"{model_py.rsquared:.3f}",
        "~5,602 bbl/day (with HAC range in scenario table)",
    ],
})
display(kpi_df)
```

:::

---

## Reproducibility & Package Citations

::: {.panel-tabset}

## R

```{r pkg-citations-r}
pkg_list <- c("readr", "dplyr", "tidyr", "ggplot2", "lubridate", "scales",
              "patchwork", "broom", "sandwich", "lmtest", "knitr")

pick_first_nonempty <- function(...) {
  vals <- list(...)
  vals <- vals[!vapply(vals, function(x) is.null(x) || length(x) == 0 || identical(x, ""), logical(1))]
  if (length(vals) == 0) "See R citation()" else as.character(vals[[1]])
}

pkg_meta <- purrr::map_dfr(pkg_list, function(p) {
  d <- utils::packageDescription(p)
  tibble(
    Package = p,
    Version = as.character(d$Version),
    Citation = pick_first_nonempty(d$`URL`, d$`BugReports`)
  )
})

kable(pkg_meta, caption = "R package versions used in this report") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Python

```{python pkg-citations-py}
import importlib

mods = ["pandas", "numpy", "matplotlib", "seaborn", "scipy", "statsmodels"]
rows = []
for m in mods:
    module = importlib.import_module(m)
    rows.append({"Package": m, "Version": getattr(module, "__version__", "unknown")})
display(pd.DataFrame(rows))
```

:::

---

## Section C — Defence Readiness Appendix

The defence evaluates how clearly each technique can be explained under pressure. The four universal prompts are answered below per technique: what it does, why it was chosen, key assumption, and immediate management action.

| Technique | What It Does | Why Chosen Here | Key Assumption to Defend | One-Line Management Action |
|---|---|---|---|---|
| EDA | Describes distribution shape, outliers, and station patterns | Needed to validate data quality before inference | Extreme low values are operational events, not input errors | Prioritize CC2/CC3 reliability remediation over incident-by-incident response |
| Visualization | Shows temporal deterioration and station concentration | Required for executive communication and trend diagnosis | Visual trends should be supported by tests (Jan-Apr trend test included) | Track monthly shortfall trajectory as a formal KPI |
| Hypothesis Testing | Tests whether differences are random vs systematic | Needed for defensible resource allocation decisions | Non-normality shifts primary inference to Kruskal-Wallis | Allocate by station-level evidence first, not zone average alone |
| Correlation | Measures association strength and direction | Screens realistic levers before regression | Correlation is associational, not causal | Treat capacity as a high-priority lever with caution on causality |
| Regression | Quantifies conditional contribution of each predictor | Converts analytics into forecastable operational impact | OLS assumptions; robust SE and validation reduce but do not remove risk | Pilot staged capacity expansion and monitor realized vs forecast band |

---

## Bonus +5 — GitHub Repository Placeholder

- **Repository URL placeholder:** `https://github.com/<your-username>/oml18-capstone-abiba`
- **Minimum publish-ready structure:**
  - `README.md` (project objective, dataset dictionary, run steps)
  - `abiba_capstone_v2.qmd`
  - `data/` (or redacted sample + data access note)
  - `requirements.txt` and R package list
- **README checklist (for bonus audit):**
  1. Business question and context
  2. Exact render command
  3. Key outputs expected after render
  4. Data access/governance caveat
  5. Contact or provenance statement

---

## References

Adi, B. (2026). *AI-powered business analytics: A practical textbook for data-driven decision making — from data fundamentals to machine learning in Python and R*. Lagos Business School / markanalytics.online. https://markanalytics.online

Allaire, J. J., Teague, C., Scheidegger, C., Xie, Y., & Dervieux, C. (2022). *Quarto* (Version 1.x) [Computer software]. https://doi.org/10.5281/zenodo.5960048

Grolemund, G., & Wickham, H. (2011). Dates and times made easy with lubridate. *Journal of Statistical Software*, *40*(3), 1–25. https://doi.org/10.18637/jss.v040.i03

Hunter, J. D. (2007). Matplotlib: A 2D graphics environment. *Computing in Science & Engineering*, *9*(3), 90–95. https://doi.org/10.1109/MCSE.2007.55

McKinney, W. (2010). Data structures for statistical computing in Python. In *Proceedings of the 9th Python in Science Conference* (pp. 56–61). https://doi.org/10.25080/Majora-92bf1922-00a

Pedersen, T. L. (2024). *patchwork: The composer of plots* (R package version 1.x). https://CRAN.R-project.org/package=patchwork

R Core Team. (2024). *R: A language and environment for statistical computing* (Version 4.x). R Foundation for Statistical Computing. https://www.R-project.org/

Seabold, S., & Perktold, J. (2010). statsmodels: Econometric and statistical modeling with Python. In *Proceedings of the 9th Python in Science Conference* (pp. 92–96). https://doi.org/10.25080/Majora-92bf1922-011

Van Rossum, G., & Drake, F. L. (2009). *Python 3 reference manual*. CreateSpace.

Waskom, M. L. (2021). seaborn: Statistical data visualization. *Journal of Open Source Software*, *6*(60), 3021. https://doi.org/10.21105/joss.03021

Wickham, H. (2016). *ggplot2: Elegant graphics for data analysis*. Springer. https://doi.org/10.1007/978-3-319-24277-4

Wickham, H., Averick, M., Bryan, J., Chang, W., McGowan, L., François, R., Grolemund, G., Hayes, A., Henry, L., Hester, J., Kuhn, M., Pedersen, T. L., Miller, E., Bache, S. M., Müller, K., Ooms, J., Robinson, D., Seidel, D. P., Spinu, V., … Yutani, H. (2019). Welcome to the tidyverse. *Journal of Open Source Software*, *4*(43), 1686. https://doi.org/10.21105/joss.01686

Wickham, H., & Bryan, J. (2024). *readr: Read rectangular text data* (R package version 2.x). https://CRAN.R-project.org/package=readr

Xie, Y. (2024). *knitr: A general-purpose package for dynamic report generation in R* (R package version 1.x). https://yihui.org/knitr/

*Dataset citation:*
Badmus, A. (2026). *OML 18 daily production and alternative evacuation dataset — January to May 2026* [Dataset]. Upstream Operations Department, OML 18. Data available on request from the author subject to organisational data-governance approval.

---

## Appendix: AI Usage Statement

GitHub Copilot (Microsoft) and ChatGPT (OpenAI) were used to accelerate structure design, code templating, and review of statistical workflow sequencing in this document. All dataset-specific cleaning rules (the zero/blank `ACTUAL` imputation logic, the 150% capacity buffer threshold derivation), test selection decisions (the choice of Kruskal-Wallis as the primary zone test over ANOVA given confirmed non-normality), threshold definitions, coefficient interpretation, and operational recommendations were independently validated by the author against the actual outputs generated from internal production and evacuation data. The partial correlation specification and the `gg_corr_tile` heatmap replacement function were written by the author to resolve a system dependency issue without degrading analytical quality. Final conclusions and business actions — including the 5,602 bbl/day recovery projection and the Zone 1 fleet-expansion recommendation — reflect the author's professional judgement and accountability as Head of Alternative Crude Oil Evacuation.
