---
title: "Crude Oil Production and Evacuation Analytics: Drivers of Zone-Level Throughput Performance"
author: "Abiba Badmus — 2026/31/E/1647"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
# ── Python configuration ─────────────────────────────────────────────────────
# Configure reticulate to use venv Python BEFORE Python chunks run
if (requireNamespace("reticulate", quietly = TRUE)) {
  venv_python <- "/Users/user/Documents/LandMan/UA High/LBA -EMBA -Nomso/.venv/bin/python"
  if (file.exists(venv_python)) {
    reticulate::use_python(venv_python, required = FALSE)
  }
}

# ── R libraries ──────────────────────────────────────────────────────────────
# Load individual packages (confirmed installed); avoids stringi/tidyverse issues
pkgs_core <- c("readr", "dplyr", "tidyr", "tibble", "purrr", "forcats",
               "ggplot2", "lubridate", "scales", "ggcorrplot",
               "patchwork", "broom", "knitr")
invisible(lapply(pkgs_core, library, character.only = TRUE))

# kableExtra: shim if system dep (textshaping/harfbuzz) not available
if (!requireNamespace("kableExtra", quietly = TRUE)) {
  kable_styling <- function(x, ...) x
} else {
  library(kableExtra)
}

knitr::opts_chunk$set(fig.width = 9, fig.height = 5, dpi = 150)
```

```{python setup-py, include=FALSE}
# ── Python libraries ──────────────────────────────────────────────────────────
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
import matplotlib.ticker as mticker
import seaborn as sns
from scipy import stats
import statsmodels.formula.api as smf
import statsmodels.api as sm
from IPython.display import display

sns.set_theme(style="whitegrid", palette="muted")
plt.rcParams["figure.dpi"] = 150
```

```{r load-r, include=FALSE}
# ── R: Load and clean data ───────────────────────────────────────────────────
parse_num <- function(x) as.numeric(gsub("[^0-9.]", "", trimws(x)))

prod_r <- read_csv("data/daily production report.csv",
                   col_types = cols(.default = col_character())) |>
  mutate(
    DATE       = dmy(DATE),
    BUDGET     = parse_num(BUDGET),
    ACTUAL     = ifelse(trimws(ACTUAL) %in% c("-", "--", "  -   ", ""), "0", ACTUAL),
    ACTUAL     = parse_num(ACTUAL),
    ZONE       = trimws(ZONE),
    FLOW_STATIONS = trimws(FLOW_STATIONS),
    shortfall  = ACTUAL - BUDGET
  )

evac_r <- read_csv("data/evacuation_report.csv",
                   col_types = cols(.default = col_character())) |>
  mutate(
    DATE                  = dmy(DATE),
    evacuation_capacity   = parse_num(evacuation_capacity),
    evacuation_utilization = parse_num(evacuation_utilization),
    ZONE                  = trimws(ZONE),
    utilization_rate      = ifelse(evacuation_capacity > 0,
                                   evacuation_utilization / evacuation_capacity,
                                   NA_real_)
  )

# Zone-level aggregated production
prod_zone_r <- prod_r |>
  group_by(DATE, ZONE) |>
  summarise(BUDGET_zone   = sum(BUDGET,    na.rm = TRUE),
            ACTUAL_zone   = sum(ACTUAL,    na.rm = TRUE),
            shortfall_zone = sum(shortfall, na.rm = TRUE),
            .groups = "drop")

# Master joined dataset
df_r <- prod_zone_r |>
  left_join(evac_r, by = c("DATE", "ZONE")) |>
  mutate(period = case_when(
    month(DATE) <= 2 ~ "Jan-Feb",
    month(DATE) <= 4 ~ "Mar-Apr",
    TRUE             ~ "May+"
  ))

cat("R merge check -> rows:", nrow(df_r),
    "| missing evacuation capacity:", sum(is.na(df_r$evacuation_capacity)),
    "\n")
```

```{python load-py, include=FALSE}
# ── Python: Load and clean data ──────────────────────────────────────────────
def parse_num(s):
    return pd.to_numeric(
        s.astype(str).str.strip().str.replace(r"[^\d.]", "", regex=True),
        errors="coerce"
    )

prod_py = pd.read_csv("data/daily production report.csv", dtype=str)
prod_py["DATE"]          = pd.to_datetime(prod_py["DATE"].str.strip(), dayfirst=True)
prod_py["BUDGET"]        = parse_num(prod_py["BUDGET"])
prod_py["ACTUAL"]        = prod_py["ACTUAL"].str.strip().replace({"-": "0", "  -   ": "0", "": "0"})
prod_py["ACTUAL"]        = parse_num(prod_py["ACTUAL"]).fillna(0)
prod_py["ZONE"]          = prod_py["ZONE"].str.strip()
prod_py["FLOW_STATIONS"] = prod_py["FLOW_STATIONS"].str.strip()
prod_py["shortfall"]     = prod_py["ACTUAL"] - prod_py["BUDGET"]

evac_py = pd.read_csv("data/evacuation_report.csv", dtype=str)
evac_py["DATE"]                   = pd.to_datetime(evac_py["DATE"].str.strip(), dayfirst=True)
evac_py["evacuation_capacity"]    = parse_num(evac_py["evacuation_capacity"])
evac_py["evacuation_utilization"] = parse_num(evac_py["evacuation_utilization"])
evac_py["ZONE"]                   = evac_py["ZONE"].str.strip()
evac_py["utilization_rate"]       = evac_py["evacuation_utilization"] / evac_py["evacuation_capacity"].replace(0, np.nan)

prod_zone_py = prod_py.groupby(["DATE", "ZONE"]).agg(
    BUDGET_zone=("BUDGET", "sum"),
    ACTUAL_zone=("ACTUAL", "sum"),
    shortfall_zone=("shortfall", "sum")
).reset_index()

df_py = prod_zone_py.merge(evac_py, on=["DATE", "ZONE"], how="left")
print(f"Python merge check -> rows: {len(df_py)} | missing evacuation capacity: {df_py['evacuation_capacity'].isna().sum()}")
```

---

## 1. Executive Summary

This analysis evaluates where daily production losses are occurring and which operational levers can recover output against plan. The business context is high-stakes: sustaining throughput toward national production obligations and an internal 2026 evacuation target of 18 million barrels. I analyzed two internal operational datasets extracted from OML 18 (Oil Mining Lease 18) systems spanning January 1 to May 5, 2026: 625 station-level production records across four active flow stations — Zone 1 (CC2 and CC3) and Zone 2 (ALK and CC1) — and 250 zone-level evacuation records.

Findings show persistent underperformance across both zones, with average daily shortfalls of -2,198 bbl in Zone 1 and -1,830 bbl in Zone 2. The zone gap is statistically significant under ANOVA (F = 4.083, p = 0.0437), although effect size is small (eta-squared = 0.0065), indicating broad system pressure rather than a single-zone failure. At station level, differences are strong (ANOVA F = 32.599, p < 0.001), with CC2 worst on average shortfall (-2,838 bbl/day). Correlation and regression both indicate evacuation capacity is a key driver of shortfall reduction (r with shortfall = -0.3609; regression coefficient = -0.5602, p < 0.001), while utilization rate is positive but not significant in the multivariate model.

Recommendation: prioritize fleet expansion on Zone 1 alternative evacuation routes (CC2 and CC3). A 10,000 bbl/day increase in effective evacuation capacity is associated with approximately 5,602 bbl/day shortfall recovery — equivalent to ~2.05 million additional barrels annually, closing approximately 11.4% of the 18 million barrel 2026 annual evacuation target gap.

---

## 2. Professional Disclosure

**Job title:** Head, Alternative Crude Oil Evacuation
**Organisation sector:** Upstream oil and gas operations - national crude production, zone-level throughput management, and evacuation logistics

### Technique 1 — Exploratory Data Analysis
As Head of Alternative Crude Oil Evacuation, every working day begins with a review of the previous day's production figures: did each flow station meet its `BUDGET` target, and what is the gap in `ACTUAL`? Before any statistical test can be run, I need to understand the distributional shape of that gap across 625 daily records spanning four flow stations (ALK, CC1, CC2, CC3) and two zones. Are shortfalls concentrated in certain stations or spread evenly? Are there extreme outlier days, perhaps caused by pipeline maintenance or force majeure, that would distort any zone-level comparison? EDA answers these questions first and ensures that every downstream test and model rests on verified, understood data rather than unexamined assumptions. This mirrors the operational health-check I conduct before presenting production performance to senior management.

**Alternative considered:** Jumping directly to regression on the combined dataset. Rejected because undetected outliers in `ACTUAL` (including the blank/zero values in early January) would silently bias the regression coefficients and produce misleading conclusions about which zones are genuinely underperforming.

**Limitation:** EDA is descriptive only. A zone that visually appears to underperform may still not do so significantly once sample variance is accounted for. Formal tests are required before drawing operational conclusions.

### Technique 2 — Data Visualization
Upstream operations leadership does not make decisions from tables of numbers. The briefings I prepare for management are built around time-series charts of daily `ACTUAL` vs `BUDGET` production, zone-level comparison charts, and capacity utilization trend lines drawn from `evacuation_capacity` and `evacuation_utilization`. The `DATE` field in both datasets enables a full January-May 2026 timeline view, while `ZONE` and `FLOW_STATIONS` provide the categorical groupings needed for comparison. Visualization allows me to spot bottleneck periods - weeks where Zone 1 (CC2/CC3) and Zone 2 (ALK/CC1) diverge - in a format that a non-technical executive can act on immediately. These charts are directly reusable in production performance briefings.

**Alternative considered:** Summary statistics tables only. Rejected because tables do not communicate temporal patterns or cross-zone divergence, precisely the patterns that matter most for evacuation scheduling decisions.

**Limitation:** Visual patterns can appear more decisive than they are. A chart showing Zone 2 consistently below budget could be driven by one outlier period. All visual findings are followed by formal hypothesis tests before policy action is recommended.

### Technique 3 — Hypothesis Testing
My organization operates four flow stations across two zones, and a persistent management question is whether Zone 1 and Zone 2 perform statistically differently against their respective `BUDGET` targets, or whether the apparent difference is within normal operational variance. If the difference is statistically significant (ANOVA or Welch t-test on production shortfall by `ZONE`), that is a resource allocation issue: it justifies redirecting evacuation capacity or maintenance attention to the underperforming zone. If it is not significant, the variability is systemic and calls for a different intervention. A p-value and effect size turn what is currently an informal management opinion into a defensible, data-backed operational directive.

**Alternative considered:** Kruskal-Wallis non-parametric test as the primary method, since daily production data can exhibit skew. This is run as a robustness check alongside parametric ANOVA.

**Limitation:** After splitting by both `ZONE` and `FLOW_STATIONS`, some station-level subgroups may have fewer observations than ideal for high-power comparisons. Zone-level comparisons are adequately powered; station-level results are interpreted with explicit power caveats.

### Technique 4 — Correlation Analysis
Before including `evacuation_capacity` and `evacuation_utilization` from the evacuation dataset as predictors in a regression model, I need to establish whether they are actually related to production performance. My operational hypothesis is that low utilization relative to capacity (`evacuation_utilization / evacuation_capacity`) should correlate with better production attainment, meaning the pipeline has headroom and is not constraining output. Correlation analysis on the merged daily dataset (joined on `DATE` and `ZONE`) confirms or refutes this hypothesis and reveals which evacuation variables are worth including as regression predictors. It also guards against multicollinearity: if `evacuation_capacity` and `evacuation_utilization` are highly correlated with each other, entering both in regression would produce unstable coefficients.

**Alternative considered:** Entering all evacuation and production variables directly into regression without prior screening. Rejected because this often produces inflated standard errors and misleading coefficient signs when predictors are correlated.

**Limitation:** Correlation measures linear association only. The relationship between evacuation utilization and production attainment may be non-linear, for example, utilization below 40% may have limited impact while utilization above 80% may sharply constrain output. This is explored visually before regression specification.

### Technique 5 — Multiple Regression
The commercial and strategic purpose of this analysis is a daily production attainment model: given a station's `BUDGET` target, its `ZONE`, its `evacuation_capacity`, and its `evacuation_utilization`, what `ACTUAL` production should be expected? A regression equation with `ACTUAL - BUDGET` shortfall as the outcome variable provides coefficient-level answers to the question every evacuation planning meeting asks: which factor, if improved by one unit, recovers the most daily barrels? If `evacuation_utilization` has a coefficient of +150 (bbl per percentage point of utilization), that directly translates into a fleet-sizing recommendation. Every coefficient in this model maps to a recoverable share of annual throughput targets.

**Alternative considered:** Random forest regression to capture non-linear interactions between zone, station, and capacity variables. Rejected because the current dataset is relatively small for stable ensemble generalization and the output would be less interpretable for management briefing and policy recommendation.

**Limitation:** The regression assumes approximately linear relationships between predictors and shortfall. If the true relationship is threshold-based (for example, output only suffers once capacity falls below a critical level), diagnostic plots will reveal this and a transformed specification can be applied.

---

## 3. Data Collection & Sampling

- **Source system:** Internal daily production and evacuation reporting systems maintained by the Upstream Operations department — OML 18 (Oil Mining Lease 18) operations.
- **Zone and station structure:** The four active flow stations are organized into two evacuation zones: Zone 1 (CC2 and CC3) and Zone 2 (ALK and CC1). Three additional planned stations are currently non-operational and excluded from this dataset. Evacuation operates via alternative (non-trunk-pipeline) routes — primarily barge and road transport — which carry distinct capacity constraints and scheduling complexity relative to primary crude export pipelines.
- **Datasets used:**
  - `daily production report.csv` - 625 daily records covering flow-station-level budget versus actual production by zone.
  - `evacuation_report.csv` - 250 daily records covering evacuation capacity and utilization by zone.
- **Collection method:** Operational data extracted directly from internal reporting systems. Records represent daily snapshots of throughput and logistical capacity across active evacuation zones.
- **Time period covered:** January 1, 2026 to May 5, 2026.
- **Sample frame:** All active flow stations and evacuation zones within the Alternative Crude Oil Evacuation program. Stations or zones with zero operational activity in the period are retained as-is.
- **Selection bias:** No deliberate sampling was applied. The dataset represents the full operational population for the period. Some `ACTUAL` values appear as zero/blank in early days; these are treated as non-operational periods, not missing data errors.
- **Anonymization approach:** Flow-station identifiers and zone codes are internal operational designations. No employee or personal data is present. No additional masking is required.
- **Ethical considerations:** Dataset is extracted from internal company systems for which I hold operational accountability. Use is limited to this academic exercise. No external-party or third-party client data are involved.
- **Organizational approval:** Internal line management approval and data-governance clearance were obtained for academic use of anonymized operational aggregates.

---

## 4. Data Description

::: {.panel-tabset}

## R

```{r eda-describe-r}
# Dataset dimensions
cat("Production records:", nrow(prod_r), "rows |", ncol(prod_r), "columns\n")
cat("Evacuation records:", nrow(evac_r), "rows |", ncol(evac_r), "columns\n")
cat("Merged zone dataset:", nrow(df_r), "rows\n\n")

# Summary statistics for key numeric variables
prod_r |>
  select(BUDGET, ACTUAL, shortfall) |>
  summary() |>
  kable(caption = "Production dataset — key numeric summaries") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r missing-r}
# Missing value audit
prod_r |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "Column", values_to = "NA_count") |>
  filter(NA_count > 0) |>
  kable(caption = "Missing values in production dataset") |>
  kable_styling(full_width = FALSE)

cat("\nNote: ACTUAL = 0 records (non-operational days):",
    sum(prod_r$ACTUAL == 0), "\n")

# Record counts by zone and station
prod_r |>
  count(ZONE, FLOW_STATIONS) |>
  kable(caption = "Record counts by Zone and Flow Station") |>
  kable_styling(full_width = FALSE)
```

## Python

```{python eda-describe-py}
print(f"Production records: {prod_py.shape[0]} rows | {prod_py.shape[1]} columns")
print(f"Evacuation records: {evac_py.shape[0]} rows | {evac_py.shape[1]} columns")
print(f"Merged zone dataset: {df_py.shape[0]} rows\n")

display(prod_py[["BUDGET", "ACTUAL", "shortfall"]].describe().round(2))
```

```{python missing-py}
na_counts = prod_py.isnull().sum()
na_counts = na_counts[na_counts > 0]
if len(na_counts):
    print("Missing values:\n", na_counts)
else:
    print("No missing values after cleaning.\n")

print(f"ACTUAL = 0 records (non-operational days): {(prod_py['ACTUAL'] == 0).sum()}")
print("\nRecord counts by Zone and Flow Station:")
display(prod_py.groupby(["ZONE", "FLOW_STATIONS"]).size().reset_index(name="count"))
```

:::

---

## 5. Technique 1 — Exploratory Data Analysis

**Business justification:** Before any inferential test, this section establishes the shape of the production shortfall distribution, identifies extreme outlier days, and reveals how performance varies across zones and flow stations. These findings directly inform which statistical tests are appropriate in later sections.

::: {.panel-tabset}

## R

```{r eda-r}
# Distribution of shortfall by zone
p1 <- ggplot(prod_r, aes(x = shortfall, fill = ZONE)) +
  geom_histogram(bins = 40, alpha = 0.7, position = "identity") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "red") +
  facet_wrap(~ZONE, ncol = 1) +
  scale_x_continuous(labels = comma) +
  labs(title = "Distribution of Daily Production Shortfall by Zone",
       subtitle = "Red line = on-target (shortfall = 0). Left of line = underperformance.",
       x = "Shortfall (bbl/day)", y = "Count") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")

# Box plot: shortfall by flow station
p2 <- ggplot(prod_r, aes(x = FLOW_STATIONS, y = shortfall, fill = ZONE)) +
  geom_boxplot(alpha = 0.7, outlier.colour = "red", outlier.size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_y_continuous(labels = comma) +
  labs(title = "Shortfall Distribution by Flow Station",
       x = "Flow Station", y = "Shortfall (bbl/day)") +
  theme_minimal(base_size = 11)

p1 / p2
```

```{r eda-skewkurt-r}
# Skewness and kurtosis of production shortfall — required distributional statistics
sk_fn <- function(x) {
  x <- x[!is.na(x)]; n <- length(x); m <- mean(x); s <- sd(x)
  sum((x - m)^3) / (n * s^3)
}
kt_fn <- function(x) {
  x <- x[!is.na(x)]; n <- length(x); m <- mean(x); s <- sd(x)
  sum((x - m)^4) / (n * s^4) - 3  # excess kurtosis
}
sk <- sk_fn(prod_r$shortfall)
kt <- kt_fn(prod_r$shortfall)
cat(sprintf("Production shortfall — Skewness: %.3f | Excess Kurtosis: %.3f\n", sk, kt))
cat(ifelse(sk < -0.5,
  "Left-skewed: underperformance days are more extreme and frequent than overperformance days.\n",
  ifelse(sk > 0.5, "Right-skewed distribution.\n", "Approximately symmetric distribution.\n")))
cat(ifelse(kt > 1,
  "Leptokurtic (heavy tails): extreme shortfall days are more frequent than normal, consistent with maintenance events and force majeure.\n",
  "Near-normal or platykurtic tails.\n"))
```

```{r eda-summary-r}
# Summary statistics by zone and station
prod_r |>
  group_by(ZONE, FLOW_STATIONS) |>
  summarise(
    n             = n(),
    mean_budget   = mean(BUDGET,    na.rm = TRUE),
    mean_actual   = mean(ACTUAL,    na.rm = TRUE),
    mean_shortfall = mean(shortfall, na.rm = TRUE),
    pct_days_under = mean(shortfall < 0) * 100,
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric), ~ round(., 1))) |>
  kable(caption = "EDA Summary: mean daily production and shortfall by zone and station",
        col.names = c("Zone","Station","n","Avg Budget","Avg Actual","Avg Shortfall","% Days Under Budget")) |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Python

```{python eda-py}
zones = prod_py["ZONE"].dropna().unique()
fig, axes = plt.subplots(1, len(zones), figsize=(6 * len(zones), 4), squeeze=False)

for ax, zone in zip(axes[0], zones):
    subset = prod_py[prod_py["ZONE"] == zone]["shortfall"]
    ax.hist(subset, bins=40, alpha=0.7, edgecolor="white")
    ax.axvline(0, color="red", linestyle="--", label="On-target")
    ax.set_title(f"Shortfall Distribution — {zone}")
    ax.set_xlabel("Shortfall (bbl/day)")
    ax.set_ylabel("Count")
    ax.legend()

plt.suptitle("Daily Production Shortfall by Zone", fontsize=12, fontweight="bold")
plt.tight_layout()
plt.show()
```

```{python eda-summary-py}
summary = prod_py.groupby(["ZONE", "FLOW_STATIONS"]).agg(
    n=("shortfall", "count"),
    mean_budget=("BUDGET", "mean"),
    mean_actual=("ACTUAL", "mean"),
    mean_shortfall=("shortfall", "mean"),
    pct_days_under=("shortfall", lambda x: (x < 0).mean() * 100)
).round(1).reset_index()
display(summary)
```

```{python eda-skewkurt-py}
from scipy.stats import skew, kurtosis as sp_kurtosis

sk_val = skew(prod_py["shortfall"].dropna())
kt_val = sp_kurtosis(prod_py["shortfall"].dropna())  # excess kurtosis by default
print(f"Production shortfall — Skewness: {sk_val:.3f} | Excess Kurtosis: {kt_val:.3f}")
print("Left-skewed: underperformance days dominate." if sk_val < -0.5 else
      ("Right-skewed distribution." if sk_val > 0.5 else "Approximately symmetric distribution."))
print("Leptokurtic (heavy tails — extreme shortfall days more frequent than normal)." if kt_val > 1
      else "Near-normal or platykurtic tails.")
```

:::

**Interpretation:** The shortfall distribution is left-skewed (negative side), meaning underperformance days are more frequent and deeper than overperformance days. Across the full sample, 86.4% of station-days fall below budget, confirming systemic production pressure rather than isolated volatility. At station level, CC2 has the largest mean deficit (-2,837.63 bbl/day), while CC3 has the highest share of under-budget days (92.8%), indicating highly persistent reliability constraints. Zone-level averages also show sustained deficits (Zone 1: -2,198.12 bbl/day; Zone 2: -1,829.56 bbl/day). Operationally, this pattern indicates that current throughput instability is chronic and concentrated in specific assets, so corrective action should prioritize persistent bottlenecks (CC2 and CC3) instead of one-off incident response.

---

## 6. Technique 2 — Data Visualization

**Business justification:** Time-series and comparative charts communicate throughput trends and zone performance to management in the format used in board briefings. The merged dataset enables visualizing how evacuation capacity tracks alongside production attainment.

::: {.panel-tabset}

## R

```{r viz-timeseries-r}
# Time series: zone-level ACTUAL vs BUDGET
ggplot(df_r, aes(x = DATE)) +
  geom_line(aes(y = ACTUAL_zone, colour = "Actual"),  linewidth = 0.8) +
  geom_line(aes(y = BUDGET_zone, colour = "Budget"), linetype = "dashed", linewidth = 0.8) +
  facet_wrap(~ZONE, ncol = 1, scales = "free_y") +
  scale_y_continuous(labels = comma) +
  scale_colour_manual(values = c("Actual" = "#2c7bb6", "Budget" = "#d7191c")) +
  labs(title    = "Daily Production: Actual vs Budget by Zone",
       subtitle = "Jan–May 2026",
       x = "Date", y = "Production (bbl/day)", colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r viz-utilization-r}
# Evacuation utilization rate over time
ggplot(evac_r, aes(x = DATE, y = utilization_rate, colour = ZONE)) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = 1/1.5, linetype = "dashed", colour = "orange",
             linewidth = 0.8) +
  annotate("text", x = min(evac_r$DATE) + 5, y = 1/1.5 + 0.01,
           label = "Min 150% capacity threshold (66.7% utilization)", hjust = 0, size = 3) +
  scale_y_continuous(labels = percent_format()) +
  labs(title   = "Evacuation Utilization Rate by Zone",
       x = "Date", y = "Utilization Rate (util / capacity)", colour = "Zone") +
  theme_minimal(base_size = 11)
```

```{r viz-heatmap-r}
# Heatmap: average shortfall by station and month
prod_r |>
  mutate(month_label = format(DATE, "%b %Y")) |>
  group_by(FLOW_STATIONS, month_label) |>
  summarise(avg_shortfall = mean(shortfall, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = month_label, y = FLOW_STATIONS, fill = avg_shortfall)) +
  geom_tile(colour = "white") +
  scale_fill_gradient2(low = "#d73027", mid = "white", high = "#1a9850",
                       midpoint = 0, labels = comma,
                       name = "Avg Shortfall") +
  labs(title = "Average Daily Shortfall Heatmap by Station and Month",
       x = "Month", y = "Flow Station") +
  theme_minimal(base_size = 11)
```

## Python

```{python viz-timeseries-py}
fig, axes = plt.subplots(2, 1, figsize=(12, 7), sharex=True)

for ax, zone in zip(axes, df_py["ZONE"].unique()):
    z = df_py[df_py["ZONE"] == zone].sort_values("DATE")
    ax.plot(z["DATE"], z["ACTUAL_zone"], label="Actual", color="#2c7bb6", linewidth=1.2)
    ax.plot(z["DATE"], z["BUDGET_zone"], label="Budget", color="#d7191c",
            linestyle="--", linewidth=1.2)
    ax.set_title(f"Actual vs Budget — {zone}")
    ax.yaxis.set_major_formatter(mticker.FuncFormatter(lambda x, _: f"{x:,.0f}"))
    ax.legend()

plt.xlabel("Date")
fig.suptitle("Daily Production: Actual vs Budget by Zone (Jan–May 2026)",
             fontsize=12, fontweight="bold")
plt.tight_layout()
plt.show()
```

```{python viz-utilization-py}
fig, ax = plt.subplots(figsize=(12, 4))
for zone, grp in evac_py.groupby("ZONE"):
    grp = grp.sort_values("DATE")
    ax.plot(grp["DATE"], grp["utilization_rate"], label=zone, linewidth=1.2)

ax.axhline(1/1.5, color="orange", linestyle="--", linewidth=1.2,
           label="Min 150% capacity threshold (66.7% util)")
ax.yaxis.set_major_formatter(mticker.PercentFormatter(xmax=1))
ax.set_title("Evacuation Utilization Rate by Zone")
ax.set_xlabel("Date")
ax.set_ylabel("Utilization Rate")
ax.legend()
plt.tight_layout()
plt.show()
```

```{python viz-heatmap-py}
heat = prod_py.copy()
heat["month_label"] = heat["DATE"].dt.to_period("M").astype(str)
pivot = heat.groupby(["FLOW_STATIONS", "month_label"])["shortfall"].mean().unstack()

fig, ax = plt.subplots(figsize=(9, 4))
sns.heatmap(pivot, cmap="RdYlGn", center=0, annot=True, fmt=".0f",
            linewidths=0.5, ax=ax, cbar_kws={"label": "Avg Shortfall (bbl/day)"})
ax.set_title("Average Daily Shortfall Heatmap by Station and Month")
ax.set_xlabel("Month")
ax.set_ylabel("Flow Station")
plt.tight_layout()
plt.show()
```

:::

**Interpretation:** The monthly shortfall trend worsens progressively from January to May: -1,026.97 (Jan), -2,092.11 (Feb), -2,229.26 (Mar), -2,693.01 (Apr), and -3,204.92 bbl/day (May), indicating chronic deterioration rather than seasonal recovery. The 66.7% utilization threshold shown on the chart is derived from the organizational policy of maintaining evacuation infrastructure at a minimum of 150% of budgeted production volume — that is, for every 100 bbl/day of planned production, the evacuation system must carry at least 150 bbl/day of capacity. A utilization rate of 1/1.5 = 66.7% is therefore the maximum acceptable operating level before that safety buffer is fully consumed; anything higher signals the system is running without adequate headroom. Both zones consistently operate well below this threshold, meeting or exceeding it on only 12.4% of observations, with monthly means between 43.9% and 59.9% from January to April — meaning evacuation capacity headroom exists but is not translating into throughput recovery. This indicates the primary constraint is upstream of the evacuation route, at the well and flow station level (CC2 and CC3 in particular), rather than in the evacuation infrastructure itself.

---

## 7. Technique 3 — Hypothesis Testing

**Business justification:** Formally tests whether Zone 1 and Zone 2 perform statistically differently against their respective budget targets — converting an operational observation into a statistically defensible basis for resource allocation decisions.

::: {.panel-tabset}

## R

```{r hyp-r}
# ── Test 1: One-way ANOVA — shortfall across zones ───────────────────────────
aov_zone <- aov(shortfall ~ ZONE, data = prod_r)
cat("=== ANOVA: Shortfall by Zone ===\n")
aov_summary <- summary(aov_zone)
print(aov_summary[[1]])

# Effect size (eta-squared) - calculate directly from ANOVA table
aov_table <- aov_summary[[1]]
ss_between <- aov_table["ZONE", "Sum Sq"]
ss_total <- sum(aov_table[, "Sum Sq"])
eta_sq <- ss_between / ss_total
cat(sprintf("\nEta-squared (effect size): %.4f\n", eta_sq))
cat("Interpretation: ", ifelse(eta_sq < 0.01, "negligible",
    ifelse(eta_sq < 0.06, "small",
    ifelse(eta_sq < 0.14, "medium", "large"))), " effect\n\n")

# ── Test 2: ANOVA — shortfall across all four flow stations ─────────────────
aov_station <- aov(shortfall ~ FLOW_STATIONS, data = prod_r)
cat("=== ANOVA: Shortfall by Flow Station ===\n")
summary(aov_station)

# Tukey HSD post-hoc if significant
tukey <- TukeyHSD(aov_station)
cat("\n=== Tukey HSD post-hoc pairwise comparisons (Flow Stations) ===\n")
print(tukey)
```

```{r hyp-assumptions-r}
# Normality check
zones_r <- sort(unique(prod_r$ZONE))
shapiro_z1 <- shapiro.test(prod_r$shortfall[prod_r$ZONE == zones_r[1]])
shapiro_z2 <- shapiro.test(prod_r$shortfall[prod_r$ZONE == zones_r[2]])
cat("Shapiro-Wilk normality test:\n")
cat(sprintf("  %s: W=%.4f, p=%.4f\n", zones_r[1], shapiro_z1$statistic, shapiro_z1$p.value))
cat(sprintf("  %s: W=%.4f, p=%.4f\n", zones_r[2], shapiro_z2$statistic, shapiro_z2$p.value))

# If normality violated, run non-parametric Kruskal-Wallis
kw <- kruskal.test(shortfall ~ ZONE, data = prod_r)
cat(sprintf("\nKruskal-Wallis (non-parametric): chi-sq=%.4f, df=%d, p=%.4f\n",
            kw$statistic, kw$parameter, kw$p.value))
```

## Python

```{python hyp-py}
from scipy.stats import f_oneway, kruskal, shapiro, tukey_hsd

zones = prod_py["ZONE"].unique()
groups = [prod_py[prod_py["ZONE"] == z]["shortfall"].dropna().values for z in zones]

# ANOVA
f_stat, p_anova = f_oneway(*groups)
print(f"=== One-way ANOVA: Shortfall by Zone ===")
print(f"F = {f_stat:.4f}, p = {p_anova:.4f}")
print("Result:", "Significant difference" if p_anova < 0.05 else "No significant difference",
      "at 5% level\n")

# Kruskal-Wallis (non-parametric robustness check)
h_stat, p_kw = kruskal(*groups)
print(f"=== Kruskal-Wallis (non-parametric): ===")
print(f"H = {h_stat:.4f}, p = {p_kw:.4f}\n")

# Normality check
print("=== Shapiro-Wilk Normality Test ===")
for z, g in zip(zones, groups):
    stat, p = shapiro(g[:5000])  # Shapiro limited to 5000 samples
    print(f"  {z}: W={stat:.4f}, p={p:.4f}")
```

```{python hyp-station-py}
stations = prod_py["FLOW_STATIONS"].unique()
station_groups = [prod_py[prod_py["FLOW_STATIONS"] == s]["shortfall"].dropna().values
                  for s in stations]

f_stat2, p_aov2 = f_oneway(*station_groups)
print(f"\n=== ANOVA: Shortfall by Flow Station ===")
print(f"F = {f_stat2:.4f}, p = {p_aov2:.4f}")

# Tukey HSD post-hoc
from statsmodels.stats.multicomp import pairwise_tukeyhsd
tukey_result = pairwise_tukeyhsd(
    endog=prod_py["shortfall"].dropna(),
    groups=prod_py.loc[prod_py["shortfall"].notna(), "FLOW_STATIONS"]
)
print(tukey_result.summary())
```

:::

**Interpretation:**
H0: Mean daily shortfall is equal across zones (and across stations for station-level ANOVA).
H1: At least one group has a different mean shortfall.

Zone-level ANOVA returns F = 4.0829 and p = 0.0437, technically rejecting H0 at the 5% level. However, the Shapiro-Wilk test confirms non-normal shortfall distributions in both zones (p < 0.05), violating the key ANOVA assumption. The Kruskal-Wallis test — which requires no normality assumption and is therefore the more appropriate primary reference for these data — returns H = 1.6784, p = 0.1951: a non-significant result. The zone-level difference should therefore be treated as **directionally suggestive but not statistically robust** once distributional properties are properly accounted for. Effect size reinforces this: eta-squared = 0.0065 means zone membership explains less than 1% of total shortfall variance. At station level, ANOVA is strongly significant (F = 32.5987, p < 0.001) and this result is robust given the larger group sizes and much stronger effect. **In operational terms, station-level targeting — specifically CC2 and CC3 within Zone 1 — is the primary evidence-based justification for resource allocation, not zone-level comparisons alone.**

---

## 8. Technique 4 — Correlation Analysis

**Business justification:** Maps the relationships between budget targets, actual production, evacuation capacity, and utilization before building the regression model. Identifies which evacuation variables have genuine independent predictive value.

::: {.panel-tabset}

## R

```{r corr-r}
# Correlation matrix on merged zone dataset
corr_vars <- df_r |>
  select(BUDGET_zone, ACTUAL_zone, shortfall_zone,
         evacuation_capacity, evacuation_utilization, utilization_rate) |>
  drop_na()

cor_matrix <- cor(corr_vars, method = "pearson")

# Use base R heatmap instead of ggcorrplot (which requires stringi)
par(mar = c(5, 4, 4, 2))
heatmap(cor_matrix,
  main = "Pearson Correlation Matrix — Production and Evacuation Variables",
  col = colorRampPalette(c("#d73027", "white", "#1a9850"))(20),
  symm = TRUE,
  breaks = seq(-1, 1, length.out = 21),
  cexRow = 0.9,
  cexCol = 0.9)
```

```{r corr-table-r}
# Top correlations with shortfall
cor_tbl <- as.data.frame(cor_matrix) |>
  rownames_to_column("Variable") |>
  select(Variable, shortfall_zone) |>
  filter(Variable != "shortfall_zone") |>
  arrange(desc(abs(shortfall_zone))) |>
  rename(Correlation_with_Shortfall = shortfall_zone)

kable(cor_tbl, digits = 4,
      caption = "Correlation of each variable with daily zone-level shortfall") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Python

```{python corr-py}
corr_cols = ["BUDGET_zone", "ACTUAL_zone", "shortfall_zone",
             "evacuation_capacity", "evacuation_utilization", "utilization_rate"]
corr_data = df_py[corr_cols].dropna()
cor_mat = corr_data.corr(method="pearson")

fig, ax = plt.subplots(figsize=(8, 6))
mask = np.triu(np.ones_like(cor_mat, dtype=bool))
sns.heatmap(cor_mat, mask=mask, annot=True, fmt=".3f", cmap="RdYlGn",
            center=0, vmin=-1, vmax=1,
            linewidths=0.5, ax=ax,
            cbar_kws={"shrink": 0.8})
ax.set_title("Pearson Correlation Matrix — Production & Evacuation Variables",
             fontsize=11, fontweight="bold")
plt.tight_layout()
plt.show()
```

```{python corr-table-py}
shortfall_corr = cor_mat["shortfall_zone"].drop("shortfall_zone").sort_values(
    key=abs, ascending=False
)
print("Correlation with shortfall_zone:\n")
print(shortfall_corr.round(4).to_string())
```

:::

**Interpretation:** The apparent top correlation with `shortfall_zone` is `BUDGET_zone` (r = -0.5964), but this figure is partially a mathematical artifact: since `shortfall_zone` is defined as `ACTUAL_zone − BUDGET_zone`, the budget term appears on both sides by construction. Higher budget targets mechanically produce more negative shortfall values in absolute terms regardless of operational factors, so this correlation cannot be read as an independent operational predictor. The genuinely actionable finding is the `evacuation_capacity` correlation (r = -0.3609, p < 0.05): zones with higher bbl/day evacuation capacity consistently record smaller production deficits, and this relationship holds independently of the budget scale effect. `ACTUAL_zone` (r = -0.2656) similarly reflects zone-scale. `utilization_rate` shows only a weak positive association (r = 0.1825), suggesting that the rate at which available capacity is used matters less than the absolute level of capacity itself — a finding that the regression in Section 9 confirms.

---

## 9. Technique 5 — Multiple Regression

**Business justification:** Quantifies the independent contribution of zone, evacuation capacity, and utilization to daily production shortfall. The resulting coefficients support concrete fleet-sizing and capacity-allocation recommendations.

::: {.panel-tabset}

## R

```{r reg-r}
# Encode ZONE as factor
df_r$ZONE_f <- factor(df_r$ZONE)

# Model: shortfall ~ ZONE + evacuation_capacity + utilization_rate
model <- lm(shortfall_zone ~ ZONE_f + evacuation_capacity + utilization_rate,
            data = df_r)

cat("=== Regression: Daily Zone Shortfall ===\n")
summary(model)
```

```{r reg-diagnostics-r}
# Coefficient table - extract directly from model to avoid stringi dependency
coef_data <- data.frame(
  term = names(coef(model)),
  estimate = coef(model),
  std.error = sqrt(diag(vcov(model))),
  row.names = NULL
)
coef_data$t.value <- coef_data$estimate / coef_data$std.error
coef_data$p.value <- 2 * pt(abs(coef_data$t.value), df = model$df.residual, lower.tail = FALSE)
coef_data$conf.low <- coef_data$estimate - 1.96 * coef_data$std.error
coef_data$conf.high <- coef_data$estimate + 1.96 * coef_data$std.error
coef_data[, -1] <- round(coef_data[, -1], 4)
kable(coef_data, caption = "Regression coefficients with 95% confidence intervals") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

# Model fit - extract directly from model summary
model_summary <- summary(model)
fit_data <- data.frame(
  r.squared = round(model_summary$r.squared, 4),
  adj.r.squared = round(model_summary$adj.r.squared, 4),
  fstatistic = round(model_summary$fstatistic[1], 4),
  p.value = round(pf(model_summary$fstatistic[1], model_summary$fstatistic[2], 
                     model_summary$fstatistic[3], lower.tail = FALSE), 4),
  AIC = round(AIC(model), 4),
  BIC = round(BIC(model), 4)
)
kable(fit_data, caption = "Model fit statistics") |>
  kable_styling(full_width = FALSE)
```

```{r reg-plots-r}
# Diagnostic plots
par(mfrow = c(2, 2))
plot(model, which = 1:4, pch = 16, cex = 0.6)
par(mfrow = c(1, 1))
```

```{r reg-assumptions-r}
# ── Variance Inflation Factors — manual, no car package required ──────────────
X_mat <- model.matrix(model)[, -1]  # drop intercept column
vif_vals <- sapply(seq_len(ncol(X_mat)), function(i) {
  r2_i <- summary(lm(X_mat[, i] ~ X_mat[, -i]))$r.squared
  1 / (1 - r2_i)
})
names(vif_vals) <- colnames(X_mat)
cat("=== Variance Inflation Factors (VIF) ===\n")
print(round(vif_vals, 3))
cat("Note: VIF > 5 = problematic multicollinearity; > 10 = severe\n\n")

# ── Durbin-Watson statistic — direct computation ───────────────────────────────
resids <- residuals(model)
dw_stat <- sum(diff(resids)^2) / sum(resids^2)
cat(sprintf("=== Durbin-Watson Statistic: %.4f ===\n", dw_stat))
cat("Reference: DW ≈ 2 = no autocorrelation | DW < 1.5 = positive autocorrelation (common in daily data)\n")
cat(ifelse(dw_stat < 1.5,
  "CAUTION: Positive autocorrelation indicated — OLS standard errors may be underestimated.\n",
  ifelse(dw_stat > 2.5,
  "CAUTION: Negative autocorrelation indicated.\n",
  "No strong autocorrelation indicated at this threshold.\n")))
```

## Python

```{python reg-py}
import statsmodels.formula.api as smf

df_reg = df_py.dropna(subset=["shortfall_zone", "evacuation_capacity",
                               "utilization_rate", "ZONE"])

# Encode ZONE
df_reg = df_reg.copy()
df_reg["ZONE_f"] = pd.Categorical(df_reg["ZONE"])

model_py = smf.ols(
    "shortfall_zone ~ C(ZONE_f) + evacuation_capacity + utilization_rate",
    data=df_reg
).fit()

print(model_py.summary())
```

```{python reg-diagnostics-py}
# Residual plot
fig, axes = plt.subplots(1, 2, figsize=(12, 4))

# Residuals vs Fitted
axes[0].scatter(model_py.fittedvalues, model_py.resid, alpha=0.5, s=15)
axes[0].axhline(0, color="red", linestyle="--")
axes[0].set_xlabel("Fitted values")
axes[0].set_ylabel("Residuals")
axes[0].set_title("Residuals vs Fitted")

# Q-Q plot
sm.qqplot(model_py.resid, line="s", ax=axes[1], alpha=0.5)
axes[1].set_title("Q-Q Plot of Residuals")

plt.tight_layout()
plt.show()

print(f"\nR-squared: {model_py.rsquared:.4f}")
print(f"Adj. R-squared: {model_py.rsquared_adj:.4f}")
print(f"F-statistic p-value: {model_py.f_pvalue:.6f}")
```

```{python reg-assumptions-py}
from statsmodels.stats.outliers_influence import variance_inflation_factor
from statsmodels.stats.stattools import durbin_watson

# Variance Inflation Factors
X_vif = df_reg[["evacuation_capacity", "utilization_rate"]].dropna().reset_index(drop=True)
vif_df = pd.DataFrame({
    "Feature": X_vif.columns,
    "VIF": [round(variance_inflation_factor(X_vif.values, i), 3) for i in range(X_vif.shape[1])]
})
print("=== Variance Inflation Factors (VIF) ===")
print(vif_df.to_string(index=False))
print("Note: VIF > 5 = problematic multicollinearity; > 10 = severe\n")

# Durbin-Watson autocorrelation test
dw_stat = durbin_watson(model_py.resid)
print(f"=== Durbin-Watson Statistic: {dw_stat:.4f} ===")
print("Reference: DW ≈ 2 = no autocorrelation | DW < 1.5 = positive autocorrelation")
if dw_stat < 1.5:
    print("CAUTION: Positive autocorrelation — OLS standard errors may be underestimated.")
elif dw_stat > 2.5:
    print("CAUTION: Negative autocorrelation detected.")
else:
    print("No strong autocorrelation indicated at this threshold.")
```

:::

**Interpretation:** The model explains 29.0% of shortfall variance (R² = 0.2904; adj. R² = 0.2813; overall p < 0.001). For operational daily data this is acceptable — the remaining 71% is attributable to unmeasured factors including maintenance events, force majeure, feed-source variability, and staffing. Zone 1 (CC2 + CC3) is the reference group; the Zone 2 (ALK + CC1) coefficient of +2,987.76 bbl/day (p < 0.001) means Zone 2 shortfall is approximately 2,988 bbl/day less severe than Zone 1 after controlling for capacity and utilization — confirming Zone 1 as the primary operational problem area. `evacuation_capacity` is the key modifiable lever: each additional 1 bbl/day of evacuation capacity is associated with a 0.5602 bbl/day reduction in shortfall (p < 0.001). Scaled to a practical intervention: a 10,000 bbl/day capacity increase on Zone 1 alternative routes is projected to recover approximately 5,602 bbl/day — equivalent to ~2.05 million additional barrels annually, closing approximately 11.4% of the 18 million barrel 2026 target. `utilization_rate` is not statistically significant (p = 0.755); once absolute capacity is controlled, the utilization rate adds no independent explanatory value. **Diagnostic assessment:** The residuals vs. fitted plot shows no strong systematic fan pattern, indicating linearity and homoscedasticity assumptions are broadly satisfied. The Q-Q plot shows mild tail deviation consistent with the non-normality confirmed in Section 7; with n ≈ 250 zone-level observations this is within acceptable range under the central limit theorem. No single observation dominates the Cook's distance plot. Coefficient interpretations and the capacity intervention projection are robust.

---

## 10. Integrated Findings

The five analyses tell a coherent story about crude oil production performance:

- **EDA** established that underperformance is persistent and widespread: 86.4% of observations are below budget, with CC2 showing the deepest average shortfall and CC3 the highest proportion of under-budget days.
- **Visualization** confirmed a worsening trajectory from January to May, with mean shortfall deepening monthly and utilization generally below the 66.7% threshold.
- **Hypothesis testing** showed a statistically significant zone difference under ANOVA (p = 0.0437), but a small effect size, while station-level differences are much stronger (p < 0.001).
- **Correlation analysis** identified evacuation capacity as a meaningful operational lever (r with shortfall = -0.3609), stronger than utilization rate.
- **Regression** quantified this relationship: after controlling for zone, each 1-unit capacity increase is associated with a 0.5602 bbl/day shortfall improvement, and the model explains 29.0% of shortfall variance.

**Recommendation:** Implement a targeted fleet expansion program on Zone 1 alternative evacuation routes (CC2 and CC3). Zone 1 currently operates at 14,000 bbl/day evacuation capacity; incrementing this toward a level that sustains utilization above 70% while retaining the mandatory 150% capacity buffer would materially reduce the throughput gap. A 10,000 bbl/day increase in effective evacuation capacity is projected to recover approximately 5,602 bbl/day in production attainment. Sustained over a full year, this translates to approximately 2.05 million additional barrels — closing approximately 11.4% of the 18 million barrel 2026 annual evacuation target gap. Fleet expansion on Zone 1 alternative routes is therefore the highest-leverage single operational intervention identifiable within this dataset.

---

## 11. Limitations & Further Work

- May 2026 is a partial month in the sample window (up to May 5 only), so month-end conclusions should be interpreted cautiously and re-validated with full-month data.
- The regression assumes linear relationships between evacuation variables and production shortfall. Threshold effects (e.g., non-linear impact of capacity falling below a critical level) should be explored with polynomial or spline terms.
- Three of the seven planned flow stations are currently non-functional and excluded from this analysis. A full production performance model should account for planned station reactivation.
- Future work: incorporate pipeline maintenance event data as an additional predictor to isolate scheduled downtime from structural underperformance.

---

## References

- Wickham, H., & Grolemund, G. (2017). *R for Data Science*. O'Reilly Media.
- McKinney, W. (2022). *Python for Data Analysis* (3rd ed.). O'Reilly Media.
- Adi, B. (2026). *AI-powered business analytics: A practical textbook for data-driven decision making - from data fundamentals to machine learning in Python and R*. Lagos Business School / markanalytics.online. https://markanalytics.online
- R Core Team. (2024). *R: A language and environment for statistical computing*. R Foundation for Statistical Computing.
- Pedersen, T. L. (2024). *patchwork: The Composer of Plots* (R package).
- Kassambara, A. (2024). *ggcorrplot: Visualization of a Correlation Matrix using ggplot2* (R package).
- Xie, Y. (2024). *knitr: A General-Purpose Package for Dynamic Report Generation in R* (R package).

---

## Appendix: AI Usage Statement

GitHub Copilot (Microsoft) and ChatGPT (OpenAI) were used to accelerate structure design, code templating, and review of statistical workflow sequencing. All dataset-specific cleaning rules, test selection, threshold definitions, coefficient interpretation, and operational recommendations were independently validated by the author against the actual outputs generated from internal production and evacuation data. Final conclusions and business actions reflect the author's professional judgement and accountability.
